# GeneTrackR

GeneTrackR is an R package for GenePred annotation processing and genomic signal track visualization.

## Main features

- Read, validate, summarize, slice, merge, and write GenePred/GenePredExt annotations.
- Read bedGraph, wig, and bigWig signal tracks.
- Query signal tracks by genomic region.
- Plot transcript, gene, and region-level gene models.
- Plot genomic signal as bar, line, area, or heatmap tracks.
- Combine signal tracks and gene model tracks in genome-browser-like layouts.

## Development

```r
install.packages(c("devtools", "roxygen2"))
devtools::document()
devtools::install()
```

## 1. First example
### 1.1 操作gtf/genepred注释文件
```{r}
################################################
ncu12 <- read_genepred(
  'E:/rensc/PHD/1404/database/norm/NC12.genepred',
  format = "genePredExt"
)

ncu12_gtf <- read_gtf(
  'E:/rensc/PHD/1404/database/norm/NC12.ribo.gtf',
  verbose = TRUE
)

ncu12_gff <- read_gff(
  'E:/rensc/PHD/1404/database/norm/NC12_features.gff',
  verbose = TRUE
)

ncu12_summary <- summary_genepred(ncu12)


retrieve_feature(ncu12, gene_id = "NCU06011")

################################################
write_feature(
  ncu12,
  file = 'E:/rensc/PHD/1404/database/norm/NC12_features.gtf',
  format = 'gtf',
  overwrite = TRUE,
  sort_output = TRUE
)

write_feature(
  ncu12_gtf,
  file = 'E:/rensc/PHD/1404/database/norm/NC12_features.genepred',
  format = 'genepredext',
  overwrite = TRUE,
  sort_output = TRUE
)

write_feature(
  ncu12_gtf,
  file = 'E:/rensc/PHD/1404/database/norm/NC12_features.gff',
  format = 'gff',
  overwrite = TRUE,
  sort_output = TRUE
)

write_feature(
  ncu12,
  file = 'E:/rensc/PHD/1404/database/norm/NC12_features.bed',
  format = 'bed',
  overwrite = TRUE,
  sort_output = TRUE
)
################################################
plot_gene_length_distribution(
  ncu12,
  feature = "all",
  group_by = "feature",
  fill_colors = c(
    gene = "#a6cee3",
    transcript = "#1f78b4",
    exon = "#b2df8a",
    cds = "#33a02c",
    utr = "#fb9a99",
    five_utr = "#e31a1c",
    three_utr = "#fdbf6f"
  )
)

################################################
plot_gene(
  ncu12_gtf,
  gene_id = "NCU06011",
  collapse = 'none',
  coordinate = 'genomic',
  label_position = 'feature'
)

plot_transcript(
  ncu12,
  transcript_id = "EAA34829",
  border_color = "black",
  coordinate = 'transcript'
)

plot_region(
  ncu12_gtf,
  chrom = "I",
  start = 9100000,
  end = 9160000,
  text_size = 14,
  label_size = 12,
  mode = 'trim',
  collapse = 'none',
  color_palette = "Paired",
  label_position = 'feature',
  label_side = "above",
  direction_mode = "end",
  label_by = 'gene'
)
```

### 1.2 操作bedgraph/bigwig信号文件
```{r}
################################################
bg1404 <- read_bwg(
  files = c(
    'E:/rensc/PHD/1404/bedgraph/mrna-rep2/KO1.mRNA2.bedgraph.gz',
    'E:/rensc/PHD/1404/bedgraph/mrna-rep2/KO2.mRNA2.bedgraph.gz',
    'E:/rensc/PHD/1404/bedgraph/mrna-rep2/WT1.mRNA2.bedgraph.gz',
    'E:/rensc/PHD/1404/bedgraph/mrna-rep2/WT2.mRNA2.bedgraph.gz'
  ),
  format = 'bedgraph',
  use_tabix = 'yes',
  #sample_names = c("KO1", "KO2", "WT1", "WT2"),
  mode = 'lazy',
  verbose = TRUE
)

################################################
system.time({
  t1 <- retrieve_bwg(
    bg1404,
    chrom = 'I',
    start = 1,
    end = 10000,
    strand = '+',
    verbose = TRUE,
    progress = TRUE
  )
})

################################################
plot_signal_gene(
  bg1404,
  annotation = ncu12,
  gene_id = "NCU02793",
  signal_y_scale = "free",
  signal_colors = c(
    "KO1" = "#4393C3",
    "KO2" = "#4393C3",
    "WT1" = "#D6604D",
    "WT2" = "#D6604D"
  ),
  plot_type = "heatmap"
)

################################################
plot_signal_transcript(
  bg1404,
  annotation = ncu12,
  transcript_id = "EAA34829",
  signal_y_scale = "fixed",
  signal_y_ticks = "pretty",
  signal_palette = "Set3",
  coordinate = "transcript",
  plot_type = "bar"
)

################################################
plot_tracks(
  ncu12,
  bg1404,
  chrom = "I",
  start = 110000,
  end = 120000,
  signal_type = 'heatmap',
  grid_linewidt = 0.1
)

ncu12_flt <- retrieve_feature(ncu12, chrom = "I", start = 110000, end = 120000)
```

### 1.3 操作bigwig信号文件
```{r}
################################################
bwg1404 <- read_bwg(
  files = c(
    'E:/rensc/PHD/1404/bigwig/bedtools/mrna3/1404_1.bigwig',
    'E:/rensc/PHD/1404/bigwig/bedtools/mrna3/1404_2.bigwig',
    'E:/rensc/PHD/1404/bigwig/bedtools/mrna3/1404_3.bigwig',
    'E:/rensc/PHD/1404/bigwig/bedtools/mrna3/87_1.bigwig',
    'E:/rensc/PHD/1404/bigwig/bedtools/mrna3/87_2.bigwig',
    'E:/rensc/PHD/1404/bigwig/bedtools/mrna3/87_3.bigwig'
  ),
  format = 'bigwig',
  mode = 'lazy'
)


t1 <- query_bwg(bwg1404, chrom = 'I', start = 1, end = 10000, strand = '+')
################################################
plot_signal_gene(
  bwg1404,
  annotation = ncu12,
  gene_id = "NCU02793",
  grid_linewidth = 0.25,
  signal_y_scale = "fixed",
  signal_transform = "sqrt",
  signal_palette = "Paired",
  plot_type = "bar"
)

plot_signal_gene(
  bwg1404,
  annotation = ncu12,
  gene_id = "NCU02793",
  signal_y_scale = "free",
  signal_colors = c(
    "1404_1" = "#4393C3",
    "1404_2" = "#4393C3",
    "1404_3" = "#4393C3",
    "87_1" = "#D6604D",
    "87_2" = "#D6604D",
    "87_3" = "#D6604D"
  ),
  plot_type = "bar"
)


plot_signal_transcript(
  bwg1404,
  annotation = ncu12,
  transcript_id = "EAA34829",
  signal_y_scale = "free",
  signal_y_ticks = "range",
  signal_palette = "Set3",
  coordinate = "transcript",
  plot_type = "bar"
)

plot_signal_region(
  bwg1404,
  annotation = ncu12,
  chrom = "I",
  start = 9100000,
  end = 9160000,
  plot_type = "bar"
)

################################################
plot_tracks(
  ncu12,
  bwg1404,
  chrom = "I",
  start = 110000,
  end = 120000
)
```


### 1.4 操作 vcf 文件

```{r}
vcf <- read_vcf('./inst/extdata/NC12_test_variants.vcf')

plot_variant(
  vcf,
  chrom = "I",
  start = 110000,
  end = 120000
)

# 按变异类型着色：
plot_variant(
  vcf,
  chrom = "I",
  start = 110000,
  end = 120000,
  color_by = "variant_type"
)

# 按 FILTER 着色：
plot_variant(
  vcf,
  chrom = "I",
  start = 110000,
  end = 120000,
  color_by = "filter"
)

# 显示 variant ID：
plot_variant(
  vcf,
  chrom = "I",
  start = 110000,
  end = 120000,
  label_by = "variant_id"
)

# 只画 SNP：
plot_variant(
  vcf,
  chrom = "I",
  start = 110000,
  end = 120000,
  variant_type = "SNP"
)

#  画完整的track图
plot_tracks(
  annotation = ncu12,
  signal = bg1404,
  variants = vcf,
  chrom = "I",
  start = 110000,
  end = 120000,
  signal_type = "bar"
)
```
















## Second example

```{r}
################################################
gmx4 <- read_genepred(
  'E:/rensc/PHD/gma/database/norm2/gmx4.genepred',
  format = "genePredExt"
)
gmx4_summary <- summary_genepred(gmx4)

search_gene(gmx4, "LOC100527567")


################################################
plot_genepred_length_distribution(
  gmx4,
  feature = "all",
  group_by = "feature",
  fill_colors = c(
    gene = "#a6cee3",
    transcript = "#1f78b4",
    exon = "#b2df8a",
    cds = "#33a02c",
    utr = "#fb9a99",
    five_utr = "#e31a1c",
    three_utr = "#fdbf6f"
  )
)

################################################
plot_gene(
  gmx4,
  gene_id = "LOC100527567",
  collapse = 'none',
  coordinate = 'genomic',
  label_position = 'feature',
  fill_colors = c(
    CDS = "#b2df8a",
    UTR = "#33a02c",
    exon = "#fb9a99"
  )
)

plot_transcript(
  gmx4,
  transcript_id = "XM_006606546.4",
  border_color = "black",
  coordinate = 'transcript',
  fill_colors = c(
    "#b2df8a",
    "#33a02c",
    "#fb9a99"
  )
)

plot_region(
  gmx4,
  chrom = "NC_038252.2",
  start = 35902888,
  end = 35923703,
  text_size = 14,
  label_size = 12,
  mode = 'trim',
  collapse = 'none',
  color_palette = "Blues",
  label_position = 'feature',
  label_side = "above",
  direction_mode = "end",
  label_by = 'gene'
)
```


```{r}
################################################
gmx_bg <- read_bwg(
  files = c(
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_01d_minus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_01d_plus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_02d_minus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_02d_plus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_04d_minus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_04d_plus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_07d_minus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_07d_plus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_10d_minus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_10d_plus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_15d_minus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_15d_plus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_20d_minus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_20d_plus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_25d_minus.rpm.sorted.bedgraph.gz',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/ribo/R_GR_in_25d_plus.rpm.sorted.bedgraph.gz'
  ),
  strand = c(
    '-',
    '+',
    '-',
    '+',
    '-',
    '+',
    '-',
    '+',
    '-',
    '+',
    '-',
    '+',
    '-',
    '+',
    '-',
    '+'
  ),
  format = 'bedgraph',
  use_tabix = 'yes',
  mode = 'lazy'
)

################################################
system.time({
  t1 <- query_bwg(
    gmx_bg,
    chrom = 'NC_038256.2',
    start = 47723683,
    end = 47953121,
    strand = '+',
    verbose = TRUE,
    progress = TRUE
  )
})

################################################
plot_signal_gene(
  gmx_bg,
  annotation = gmx4,
  sample_groups = data.frame(
    sample_id = c(
      'R_GR_in_01d_minus.rpm.sorted',
      'R_GR_in_01d_plus.rpm.sorted',
      'R_GR_in_02d_minus.rpm.sorted',
      'R_GR_in_02d_plus.rpm.sorted',
      'R_GR_in_04d_minus.rpm.sorted',
      'R_GR_in_04d_plus.rpm.sorted',
      'R_GR_in_07d_minus.rpm.sorted',
      'R_GR_in_07d_plus.rpm.sorted',
      'R_GR_in_10d_minus.rpm.sorted',
      'R_GR_in_10d_plus.rpm.sorted',
      'R_GR_in_15d_minus.rpm.sorted',
      'R_GR_in_15d_plus.rpm.sorted',
      'R_GR_in_20d_minus.rpm.sorted',
      'R_GR_in_20d_plus.rpm.sorted',
      'R_GR_in_25d_minus.rpm.sorted',
      'R_GR_in_25d_plus.rpm.sorted'
    ),
    group = c(
      "minus",
      "plus",
      "minus",
      "plus",
      "minus",
      "plus",
      "minus",
      "plus",
      "minus",
      "plus",
      "minus",
      "plus",
      "minus",
      "plus",
      "minus",
      "plus"
    )
    # group = c(
    #   "01d_minus",
    #   "01d_plus",
    #   "02d_minus",
    #   "02d_plus",
    #   "04d_minus",
    #   "04d_plus",
    #   "07d_minus",
    #   "07d_plus",
    #   "10d_minus",
    #   "10d_plus",
    #   "15d_minus",
    #   "15d_plus",
    #   "20d_minus",
    #   "20d_plus",
    #   "25d_minus",
    #   "25d_plus"
    # )
  ),
  bin_size = 10,
  signal_color_by = c("sample", "group"),
  strand = "both",
  gene_id = "LOC100527567",
  signal_y_scale = "fixed",
  signal_y_ticks = "pretty",
  signal_transform = "sqrt",
  signal_palette = "RdBu",
  plot_type = "bar"
)

################################################
plot_signal_transcript(
  gmx_bg,
  annotation = gmx4,
  bin_size = 10,
  transcript_id = "XM_006606546.4",
  signal_y_scale = "fixed",
  signal_y_ticks = "pretty",
  signal_palette = "RdBu",
  coordinate = "transcript",
  plot_type = "bar"
)

################################################
plot_tracks(
  gmx4,
  gmx_bg,
  chrom = "NC_038256.2",
  start = 47823683,
  end = 47853121
)
```


```{r}
################################################
gmx_bwg <- read_bwg(
  files = c(
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/mrna/m_GR_in_01d.rpm.sort.bigwig',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/mrna/m_GR_in_02d.rpm.sort.bigwig',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/mrna/m_GR_in_04d.rpm.sort.bigwig',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/mrna/m_GR_in_07d.rpm.sort.bigwig',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/mrna/m_GR_in_10d.rpm.sort.bigwig',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/mrna/m_GR_in_15d.rpm.sort.bigwig',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/mrna/m_GR_in_20d.rpm.sort.bigwig',
    'E:/rensc/PHD/usda110/summary-files2/8-gma_bedgraph/mrna/m_GR_in_25d.rpm.sort.bigwig'
  ),
  format = 'bigwig',
  mode = 'lazy'
)


t2 <- query_bwg(
  gmx_bwg,
  chrom = 'NC_038256.2',
  start = 47723683,
  end = 47953121,
  strand = '+',
  verbose = TRUE,
  progress = TRUE
)

################################################
plot_signal_gene(
  gmx_bwg,
  annotation = gmx4,
  gene_id = "LOC100817175",
  grid_linewidth = 0.25,
  signal_y_scale = "fixed",
  signal_transform = "sqrt",
  signal_palette = "Paired",
  plot_type = "bar"
)

plot_signal_gene(
  gmx_bwg,
  annotation = gmx4,
  gene_id = "LOC100817175",
  signal_y_scale = "free",
  plot_type = "bar"
)


plot_signal_transcript(
  gmx_bwg,
  annotation = gmx4,
  transcript_id = "XM_006606546.4",
  signal_y_scale = "free",
  signal_y_ticks = "range",
  signal_palette = "Set3",
  coordinate = "genomic",
  plot_type = "bar"
)

plot_signal_region(
  gmx_bwg,
  annotation = gmx4,
  chrom = "NC_038256.2",
  start = 47823683,
  end = 47853121,
  plot_type = "bar"
)

plot_tracks(
  gmx4,
  gmx_bg,
  chrom = "NC_038256.2",
  start = 47823683,
  end = 47853121
)

################################################
plot_tracks(
  gmx_bwg,
  gmx4,
  chrom = "NC_038256.2",
  start = 47823683,
  end = 47853121
)
```
