```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12.5,
  fig.height = 8,
  dpi = 300,
  fig.retina = 1,
  out.width = "100%",
  fig.align = "center"
)
```

`plot_tracks()` builds a genome-browser-like view in which gene models, quantitative genomic signals, interval features, and variants share one genomic x-axis. It is intended for **integrated locus inspection**. Use the dedicated `plot_gene()`, `plot_signal_*()`, `plot_feature_track()`, or `plot_variant()` functions when one track needs detailed track-specific annotation.

The browser workflow is:

1. prepare annotation plus optional signal, feature, and variant objects;
2. select exactly one locus by gene, transcript, or explicit genomic region;
3. choose which track classes should be displayed;
4. control sample selection, strand, panel layout, heights, and styling;
5. keep regional data access lazy or targeted for large datasets.

## Prepare the track objects

The bundled deterministic example contains all four track classes needed for an integrated locus view.

```{r prepare-inputs}
library(GeneTrackR)

gp <- read_genepred(
  system.file(
    "extdata",
    "gtr_demo.genePredExt",
    package = "GeneTrackR"
  ),
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)

signal_all <- read_bwg(
  system.file(
    "extdata",
    c(
      "gtr_demo_rnaseq_plus.bedgraph",
      "gtr_demo_rnaseq_minus.bedgraph",
      "gtr_demo_riboseq_plus.bedgraph",
      "gtr_demo_riboseq_minus.bedgraph"
    ),
    package = "GeneTrackR"
  ),
  format = "bedgraph",
  sample_names = c(
    "RNA_seq_plus",
    "RNA_seq_minus",
    "Ribo_seq_plus",
    "Ribo_seq_minus"
  ),
  strand = c("+", "-", "+", "-"),
  mode = "memory",
  verbose = FALSE
)

features <- read_bed(
  system.file(
    "extdata",
    "gtr_demo_features.bed",
    package = "GeneTrackR"
  ),
  verbose = FALSE,
  progress = FALSE
)

variants <- read_vcf(
  system.file(
    "extdata",
    "gtr_demo_variants.vcf",
    package = "GeneTrackR"
  ),
  mode = "memory",
  keep_genotype = TRUE,
  verbose = FALSE,
  progress = FALSE
)
```

The four inputs have distinct roles:

| Argument | Object | Role |
|---|---|---|
| `annotation` | `GenePred` / gene-model `Feature` | Defines genes, transcripts, exons, and the structural coordinate frame. |
| `signal` | `BwgTrack` | Adds quantitative RNA-seq, Ribo-seq, or other continuous signals. |
| `features` | `FeatureTrack` or named list | Adds interval annotations such as promoters, QTLs, or candidate regions. |
| `variants` | `VariantTrack` or named list | Adds VCF-derived variant positions. |

Only `annotation` is required.

## Choose exactly one genomic locator

Each `plot_tracks()` call accepts **one and only one** locus definition:

- `gene_id`;
- `transcript_id`;
- complete `chrom` + `start` + `end` coordinates.

Do not combine these locator styles in the same call. Gene and transcript locators are resolved from `annotation`; explicit coordinates are appropriate for multi-gene or externally defined candidate intervals.

## Gene-centered view

Start with the structural gene model alone.

```{r gene-only}
plot_tracks(
  annotation = gp,
  gene_id = "GeneA",
  direction_mode = "transcript",
  plot_theme = "classic",
  text_size = 9
)
```

Add strand-specific RNA-seq and Ribo-seq signals over the same gene locus.

```{r gene-with-signal}
plot_tracks(
  annotation = gp,
  signal = signal_all,
  gene_id = "GeneA",
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  signal_palette = "Paired",
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  heights = c(
    signal = 4,
    gene = 1,
    feature = 0.8,
    variant = 0.7
  ),
  plot_theme = "classic",
  text_size = 9
)
```

Unlike the gene-specific signal helpers, `plot_tracks()` does not use `strand = "auto"`. Select `+`, `-`, `both`, or `ignore` explicitly.

## Transcript-centered view and Ribo-seq frames

A transcript locator allows GeneTrackR to associate Ribo-seq signal with one transcript model while keeping every browser panel aligned in **genomic coordinates**.

```{r transcript-browser}
plot_tracks(
  annotation = gp,
  signal = signal_all,
  transcript_id = "TxA1",
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  ribo_signal_type = "auto",
  frame_palette = "Paired",
  signal_y_scale = "free",
  collapse = "none",
  direction_mode = "transcript",
  heights = c(
    signal = 4,
    gene = 1,
    feature = 0.8,
    variant = 0.7
  ),
  plot_theme = "classic",
  text_size = 9
)
```

With `ribo_signal_type = "auto"`, samples whose names look like Ribo-seq/RPF/footprint tracks are rendered in frame mode when a transcript is supplied. RNA-seq remains a standard genomic signal panel. Without a transcript locator, frame rendering is not available because one transcript model is required to define the reading frame.

## Explicit multi-gene region

Use genomic coordinates for a multi-gene candidate region or an interval defined outside the annotation.

```{r region-browser}
browser_chrom <- "chr1"
browser_start <- 12339001
browser_end <- 12374500

plot_tracks(
  annotation = gp,
  signal = signal_all,
  chrom = browser_chrom,
  start = browser_start,
  end = browser_end,
  samples = c(
    "RNA_seq_plus",
    "RNA_seq_minus",
    "Ribo_seq_plus",
    "Ribo_seq_minus"
  ),
  strand = "both",
  signal_type = "bar",
  signal_palette = "Paired",
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  collapse = "none",
  direction_mode = "gene",
  heights = c(
    signal = 5,
    gene = 1.2,
    feature = 0.8,
    variant = 0.7
  ),
  plot_theme = "classic",
  text_size = 9
)
```

Use `signal_y_scale = "free"` when assays have substantially different ranges. Use `fixed` only when a shared quantitative y-axis is biologically meaningful.

## Add feature and variant panels

`FeatureTrack` and `VariantTrack` objects are region-filtered internally to the selected browser interval.

```{r complete-browser}
plot_tracks(
  annotation = gp,
  signal = signal_all,
  features = features,
  variants = variants,
  chrom = browser_chrom,
  start = browser_start,
  end = browser_end,
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  signal_palette = "Paired",
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  gene_palette = "Paired",
  feature_color_by = "auto",
  feature_max_legend_levels = 5,
  variant_colors = c(
    SNP = "#1F78B4",
    INS = "#33A02C",
    DEL = "#E31A1C"
  ),
  direction_mode = "gene",
  heights = c(
    signal = 4,
    gene = 1.2,
    feature = 0.9,
    variant = 0.8
  ),
  plot_theme = "classic",
  text_size = 9
)
```

The integrated feature panel deliberately uses compact labeling. `feature_color_by = "auto"` attempts to select an informative low-cardinality field and `feature_max_legend_levels` limits legend complexity. Use `plot_feature_track()` when individual feature labels are required.

Integrated variant panels also hide individual labels to keep dense loci readable. Use `plot_variant()` for detailed variant labeling, then return to `plot_tracks()` for the compact overview.

Named color vectors are useful when stable biological categories should keep the same color across loci.

## Multiple independent feature or variant tracks

`features` and `variants` may be named lists. Each entry becomes an independent browser panel.

```{r multiple-track-lists, eval=FALSE}
feature_tracks <- list(
  Regulatory = regulatory_features,
  QTL = qtl_features
)

variant_tracks <- list(
  Natural = natural_variants,
  Candidate = candidate_variants
)

plot_tracks(
  annotation = gp,
  features = feature_tracks,
  variants = variant_tracks,
  chrom = browser_chrom,
  start = browser_start,
  end = browser_end,
  text_size = 9
)
```

The named `feature` and `variant` entries in `heights` are reused for every panel of that track class.

## Layout and panel heights

The default `layout = "signal_top"` orders available panels as:

1. signal;
2. feature track(s);
3. variant track(s);
4. gene model.

`layout = "gene_top"` moves the gene model above the other panels. The `heights` vector controls relative vertical space rather than absolute figure size.

```{r styled-browser}
ld_highlight <- data.frame(
  start = 12342620,
  end = 12343180
)

browser_figure <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  features = features,
  variants = variants,
  chrom = browser_chrom,
  start = browser_start,
  end = browser_end,
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  signal_transform = "sqrt",
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  signal_alpha = 0.85,
  signal_bar_width = 0.9,
  feature_color_by = "auto",
  feature_max_legend_levels = 5,
  highlight = ld_highlight,
  layout = "gene_top",
  heights = c(
    signal = 4,
    gene = 1.5,
    feature = 0.9,
    variant = 0.8
  ),
  direction_mode = "gene",
  plot_theme = "classic",
  show_panel_border = FALSE,
  text_size = 9
)

browser_figure
```

`highlight` uses genomic coordinates and is currently applied to the signal and gene-model panels. It is useful for marking a candidate interval while retaining the shared x-axis.

## Styling responsibilities

The integrated browser exposes track-specific controls without forcing every track to share one color scale.

| Track | Main controls |
|---|---|
| Signal | `signal_palette`, `signal_colors`, `signal_transform`, `signal_y_scale`, `signal_alpha`, `signal_bar_width` |
| Gene model | `gene_palette`, `gene_colors`, `gene_border_color`, `cds_height`, `utr_height`, `direction_mode`, `label_position`, `label_by` |
| Feature | `feature_color_by`, `feature_palette`, `feature_colors`, `feature_border_color`, `feature_max_legend_levels` |
| Variant | `variant_palette`, `variant_colors` |

The integrated view intentionally prioritizes alignment and compactness over exhaustive labels. Detailed interpretation should use the dedicated track-specific plotting articles first.

## Large-dataset workflow

For real whole-genome datasets, keep the browser query regional:

- use lazy `BwgTrack` access for indexed bigWig signal when possible;
- use bgzip-compressed VCF plus `.tbi` / `.csi` and `read_vcf(mode = "lazy")` for regional variants;
- preselect the signal samples needed for the biological question;
- avoid chromosome-scale rendering of dense base-level Ribo-seq or variant tracks;
- use dedicated track functions when one data type needs detailed labels or diagnostics.

`plot_tracks()` returns the assembled ggplot/patchwork figure. Figure export is handled by the standard GeneTrackR export workflow rather than duplicated here.
