```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12.5,
  fig.height = 8,
  dpi = 300,
  fig.retina = 1,
  out.width = "100%",
  fig.align = "center"
)
```

This is the production-oriented GeneTrackR workflow for a candidate gene or locus. The shorter **GeneTrackR core workflow** in Getting Started demonstrates the central package objects with fewer branches; this page additionally integrates signal tracks, interval features, browser inspection, explicit QC gates, evidence synthesis, reproducible export, and large-data adaptations. Individual Articles remain the detailed reference for module-specific parameters and interpretation limits.

The central rule is to read each source dataset once, keep the unmodified source objects, and derive locus-specific objects from them. The same annotation, VCF, signal, phenotype, region definition, and sample identifiers should then flow through downstream steps rather than being silently rebuilt with slightly different boundaries or sample sets.

The demo workflow uses `GeneA` and follows ten steps:

1. define and read the input files;
2. validate and summarize the core objects;
3. inspect the locus with annotation, signal, features, and variants;
4. retrieve the gene-body variants used for haplotype analysis;
5. construct and visualize GeneA haplotypes;
6. test haplotype and single-variant phenotype associations;
7. inspect the regional LD gradient;
8. refine genetically similar haplotypes using phenotype evidence;
9. rank regional variant effects and integrate the evidence;
10. export the results and adapt the workflow to large real datasets.

### Step 1. Define and read the inputs

Use one explicit set of source files at the start of the analysis:

```{r}
library(GeneTrackR)

gp_file <- system.file(
  "extdata",
  "gtr_demo.genePredExt",
  package = "GeneTrackR"
)

feature_file <- system.file(
  "extdata",
  "gtr_demo_features.bed",
  package = "GeneTrackR"
)

vcf_file <- system.file(
  "extdata",
  "gtr_demo_variants.vcf",
  package = "GeneTrackR"
)

pheno_file <- system.file(
  "extdata",
  "gtr_demo_pheno.tsv",
  package = "GeneTrackR"
)

signal_files <- system.file(
  "extdata",
  c(
    "gtr_demo_rnaseq_plus.bedgraph",
    "gtr_demo_rnaseq_minus.bedgraph",
    "gtr_demo_riboseq_plus.bedgraph",
    "gtr_demo_riboseq_minus.bedgraph"
  ),
  package = "GeneTrackR"
)
```

Read the objects once:

```{r}
gp <- read_genepred(
  gp_file,
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)

features <- read_bed(
  feature_file,
  verbose = FALSE,
  progress = FALSE
)

vcf <- read_vcf(
  vcf_file,
  mode = "memory",
  keep_genotype = TRUE,
  verbose = FALSE,
  progress = FALSE
)

pheno <- read_pheno(
  pheno_file,
  verbose = FALSE,
  progress = FALSE
)

signal_all <- read_bwg(
  signal_files,
  format = "bedgraph",
  sample_names = c(
    "RNA_seq_plus",
    "RNA_seq_minus",
    "Ribo_seq_plus",
    "Ribo_seq_minus"
  ),
  strand = c("+", "-", "+", "-"),
  mode = "memory",
  verbose = FALSE
)
```

For real projects, keep these source objects unchanged and create filtered/retrieved objects for each downstream question. This makes it clear which transformations were applied at each step.

### Step 2. Validate and summarize the core objects

Inspect the annotation and VCF before defining haplotypes or LD regions:

```{r}
validate_feature(gp)
validate_vcf(vcf)

summary_feature(gp)
summary_vcf(vcf)
summary_pheno(pheno)
summary_bwg(signal_all)
```

The most important consistency checks are:

- chromosome names agree among annotation, variants, and signal tracks;
- VCF genotype columns are retained for LD and haplotype analysis;
- phenotype sample IDs overlap the VCF/haplotype sample IDs;
- coordinate conventions have already been normalized by the GeneTrackR readers.

The demo phenotype order is intentionally different from the VCF sample order. Downstream phenotype functions align records by `sample_id`, not by row position.

### Step 3. Inspect the locus in a browser-like view

Start by viewing the candidate gene in its genomic context. `plot_tracks()` can combine the gene model, RNA/Ribo signal, BED-like features, and VCF variants:

```{r}
browser_geneA <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  features = features,
  variants = vcf,
  gene_id = "GeneA",
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  feature_color_by = "auto",
  feature_max_legend_levels = 5,
  direction_mode = "transcript",
  text_size = 9
)

browser_geneA
```

All general plotting palettes default to `Paired`; `plot_ld_block()` is the deliberate exception and defaults to the sequential `Reds` palette. For transcript-centered Ribo-seq inspection, supply `transcript_id` to `plot_tracks()` or use `plot_signal_transcript()` so reading-frame information can be displayed explicitly.

### Step 4. Retrieve the gene-body variants

Inspect the exact variants that will define the GeneA haplotypes:

```{r}
geneA_variants <- retrieve_vcf(
  vcf,
  annotation = gp,
  gene_id = "GeneA",
  as = "data.table",
  verbose = FALSE,
  progress = FALSE
)

geneA_variants[, .(
  variant_id,
  chrom,
  pos,
  ref,
  alt,
  variant_type
)]
```

The canonical GeneA gene body contains 11 demo variants. Do not silently add flanking variants to the primary haplotype definition: the upstream demo variant contains missing/heterogeneous genotypes and intentionally changes the resulting haplotype structure.

### Step 5. Construct and visualize the GeneA haplotypes

Use the gene body and allele-string representation for the main biological haplotype display:

```{r}
hap_gene <- hap_gene_variant(
  vcf,
  annotation = gp,
  gene_id = "GeneA",
  genotype_mode = "string"
)

hap_gene$haplotypes[, c(
  "hap_id",
  "sample_n",
  "samples"
), with = FALSE]
```

The deterministic demo contains four balanced haplotypes with nine samples each.

Plot their variant patterns together with the gene structure:

```{r}
hap_gene_figure <- plot_hap_variant(
  hap_gene,
  annotation = gp,
  min_hap_samples = 3,
  variant_label = "pos",
  show_reference_row = TRUE,
  text_size = 9
)

hap_gene_figure
```

Keep the full `HapVariant` object. Its sample assignments and genotype tables are the input for phenotype association, refinement, and variant-effect analysis.

### Step 6. Test phenotype associations

Test the original GeneA haplotypes against a trait with a designed haplotype effect:

```{r}
hap_seed <- plot_hap_pheno(
  hap_gene,
  phenotype = pheno,
  traits = "seed_weight",
  min_hap_samples = 3,
  test_method = "t.test",
  p_adjust = "BH",
  p_value_type = "adjusted",
  text_size = 9
)

hap_seed$figure
hap_seed$pvalue
```

For a specific natural variant, use the VCF genotype directly. The demo `varA03` separates the two major GeneA genotype/phenotype clusters for `protein_content`:

```{r}
varA03_protein <- plot_variant_pheno(
  variant = vcf,
  phenotype = pheno,
  variant_id = "varA03",
  traits = "protein_content",
  genotype_mode = "code",
  min_group_samples = 3,
  p_adjust = "BH",
  p_value_type = "adjusted",
  text_size = 9
)

varA03_protein$figure
```

These functions return structured result objects. Retain `$pvalue`, `$summary`, and `$plot_data` together with `$figure` rather than saving only the visual result.

### Step 7. Inspect the regional LD gradient

The primary demo LD interval contains 12 SNPs. Six form a perfect high-LD core and the remaining variants progressively decorrelate, producing high, intermediate, and low `r2` values:

```{r}
ld_gradient <- compute_ld_block(
  vcf,
  chrom = "chr1",
  start = 12342620,
  end = 12355500,
  variant_type = "snp",
  method = "r2",
  verbose = FALSE
)

ld_gradient <- plot_ld_block(
  ld_gradient,
  annotation = gp,
  show_region = TRUE,
  connect_region = TRUE,
  show_variant_marker = TRUE,
  show_variant_labels = FALSE,
  font = 9
)

ld_gradient$figure
```

Inspect the numerical result as well as the heatmap:

```{r}
ld_gradient$data[, .(
  variant_i,
  variant_j,
  distance_bp,
  r2
)]
```

GeneTrackR calculates pairwise LD; it does not automatically declare a biological block boundary from a hard-coded threshold. Define the downstream interval using the lead variant, LD pattern, gene context, and the biological question.

### Step 8. Refine haplotypes with phenotype evidence

The GeneA genotype structure contains two close haplotype pairs: `Hap1/Hap2` and `Hap3/Hap4`. Their `protein_content` values follow the same two-cluster structure, making this trait appropriate for phenotype-guided refinement:

```{r}
refined_protein <- refine_haplotype(
  hap_gene,
  phenotype = pheno,
  traits = "protein_content",
  min_hap_samples = 3,
  test_method = "t.test",
  p_adjust = "BH",
  alpha = 0.05,
  effect_threshold = 0.5
)

refined_protein$haplotype_map
refined_protein$refined_haplotypes
```

The expected deterministic grouping is:

```text
Hap1 + Hap2 -> refined group 1
Hap3 + Hap4 -> refined group 2
```

Inspect the decision evidence rather than reporting only the final group number:

```{r}
refined_protein$pairwise_test
refined_protein$phenotype_summary
```

A non-significant pairwise test alone is not proof of biological equivalence. `effect_threshold` adds an explicit phenotype-distance requirement and should be chosen in the context of the trait scale and experimental precision.

### Step 9. Rank variant effects and integrate the evidence

For interpretable two-group effect direction, rebuild the same GeneA region in binary REF/ALT-presence mode:

```{r}
hap_effect <- hap_gene_variant(
  vcf,
  annotation = gp,
  gene_id = "GeneA",
  genotype_mode = "code"
)

effect_protein <- plot_variant_effect(
  hap_effect,
  phenotype = pheno,
  traits = "protein_content",
  min_group_samples = 3,
  test_method = "t.test",
  p_adjust = "BH",
  effect_type = "absolute",
  top_n = 10,
  variant_label = "variant_id",
  x_axis = "position",
  text_size = 9
)

effect_protein$figure
```

Rank the numerical results directly:

```{r}
protein_candidates <- effect_protein$effect[
  trait == "protein_content"
][order(-abs_effect, p_adj)]

protein_candidates[, .(
  variant_id,
  effect,
  abs_effect,
  p_adj
)]
```

Several GeneA variants share the same large effect because they encode the same `Hap1/Hap2` versus `Hap3/Hap4` genotype partition. A large phenotype effect therefore identifies an associated **variant set**, not necessarily a unique causal variant.

The preferred interpretation combines all preceding evidence:

```text
variant effect
    + pairwise LD
    + gene/feature annotation
    + single-variant phenotype association
    + haplotype structure
    + phenotype-guided refinement
    -> candidate variants for biological validation
```

### Step 10. Export results and scale the workflow to real datasets

At the end of the analysis, retain the main objects together so the result can be reopened without reconstructing the workflow:

```{r}
analysis_bundle <- list(
  annotation = gp,
  features = features,
  haplotype = hap_gene,
  haplotype_pheno = hap_seed,
  variant_pheno = varA03_protein,
  ld = ld_gradient,
  refined = refined_protein,
  variant_effect = effect_protein
)

saveRDS(
  analysis_bundle,
  file = file.path(tempdir(), "GeneA.GeneTrackR_workflow.rds")
)
```

Use the export module for a complete project directory containing standard genomic tracks, PDF/PNG figures, tabular results, optional Excel workbooks, complete RDS objects, and session information.

For large real-world datasets, adapt the same workflow rather than changing its logic:

- read bgzip-compressed indexed VCF files with `read_vcf(mode = "lazy")` and always query a defined genomic interval before LD or haplotype analysis;
- use lazy bigWig access when genome-wide signal should not be loaded into memory;
- select only relevant signal samples before plotting;
- use `bin_size`, `heatmap_bin_size`, or `heatmap_max_bins` for long/high-density signal regions;
- prefer `hap_gene_variant()` and `hap_region_variant()` for new code; `hap_variant()` remains a compatibility wrapper;
- remember that direct plotting functions return ggplot/patchwork figures, whereas phenotype/variant-effect results and `LDTrack` store figures inside structured result objects.

The complete recommended chain is therefore:

```text
annotation + signal + VCF + phenotype
    -> locus inspection
    -> gene/region variant definition
    -> haplotype construction
    -> phenotype association
    -> LD interpretation
    -> phenotype-guided refinement
    -> variant-effect prioritization
    -> biological candidate selection
    -> reproducible export
```
