```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 7.5,
  dpi = 300,
  fig.retina = 1,
  out.width = "100%",
  fig.align = "center"
)
```

If you are new to the package, read **Getting started with GeneTrackR** first, then **GeneTrackR deterministic example data** for the shared file inventory, coordinate conventions, and designed truth. This article assumes that installation and the demo data model are already familiar. It is intentionally compact; use **Complete candidate-locus workflow** for the production-oriented path that also integrates signal tracks, interval features, browser inspection, reproducible export, and large-data guidance.

This vignette uses the single deterministic `gtr_demo_*` dataset shipped with GeneTrackR. The same annotation, VCF, phenotype table, and designed truth are shared across annotation, variant, haplotype, LD, refinement, and variant-effect workflows.

## Load the canonical demo inputs

```{r load-data}
library(GeneTrackR)

gp_file <- system.file("extdata", "gtr_demo.genePredExt", package = "GeneTrackR")
gtf_file <- system.file("extdata", "gtr_demo.gtf", package = "GeneTrackR")
gff_file <- system.file("extdata", "gtr_demo.gff3", package = "GeneTrackR")
bed_file <- system.file("extdata", "gtr_demo_features.bed", package = "GeneTrackR")
vcf_file <- system.file("extdata", "gtr_demo_variants.vcf", package = "GeneTrackR")
pheno_file <- system.file("extdata", "gtr_demo_pheno.tsv", package = "GeneTrackR")

gp <- read_genepred(gp_file, format = "genePredExt", verbose = FALSE)
gtf <- read_gtf(gtf_file, verbose = FALSE, progress = FALSE)
gff <- read_gff(gff_file, verbose = FALSE, progress = FALSE)
features <- read_bed(bed_file, verbose = FALSE, progress = FALSE)
vcf <- read_vcf(vcf_file, mode = "memory", verbose = FALSE)
pheno <- read_pheno(pheno_file, verbose = FALSE)
```

The three annotation formats describe the same canonical 20-gene, 24-transcript model. The VCF contains 56 designed variants in 36 samples. Phenotype row order is intentionally different from VCF sample order, so downstream analyses must align by `sample_id`.

## Object contracts used in the workflow

The workflow deliberately passes package objects between functions rather than rebuilding intermediate tables. `gp` is a `GenePred`, `features` is a `FeatureTrack`, `vcf` is a `VariantTrack`, and `pheno` is a `data.table`. The `VariantTrack` feeds variant-phenotype and LD analyses; `hap_gene_variant()` creates a `HapVariant` for haplotype plots, haplotype-phenotype analysis, refinement, and variant-effect analysis; `refine_haplotype()` creates a `HapRefined` for refined plots.

Plotting functions have two return conventions in this vignette:

- gene/track/haplotype-variant plots return a ggplot/patchwork figure directly;
- phenotype and variant-effect plots return result objects with a `$figure` field plus analysis tables;
- `plot_ld_block()` returns the updated `LDTrack` by default and stores its figure in `$figure`.

## Annotation and feature retrieval

`GeneA` is a positive-strand protein-coding gene with two transcripts. `GeneB` is a negative-strand two-transcript gene.

```{r annotation-retrieval}
genea <- retrieve_feature(gp, gene_id = "GeneA")
geneb <- retrieve_feature(gp, gene_id = "GeneB")

genea$genes
geneb$genes
```

The BED file contains promoters, enhancers, candidate regions, a QTL interval, repeats, and conserved regions around the designed loci.

```{r feature-table}
head(features$data)
```

## Variant retrieval

The main GeneA interval contains SNPs, insertions, deletions, a high-LD block, and a deliberately incomplete upstream variant.

```{r variant-retrieval}
genea_variants <- retrieve_vcf(
  vcf,
  chrom = "chr1",
  start = 12339700,
  end = 12352000,
  as = "VariantTrack",
  verbose = FALSE
)

genea_variants
genea_variants$data[, c("variant_id", "chrom", "pos", "ref", "alt"), with = FALSE]
```

`retrieve_vcf()` returns a `data.table` by default; `as = "VariantTrack"` is set here because the subset is retained as a package object.

## GeneA haplotypes

The GeneA gene body was designed to produce four balanced genotype-defined groups of nine samples each.

```{r haplotype}
hap <- hap_gene_variant(
  vcf,
  annotation = gp,
  gene_id = "GeneA",
  genotype_mode = "string",
  min_variant_number = 1
)

hap$haplotypes
```

```{r haplotype-plot}
hap_figure <- plot_hap_variant(
  hap,
  annotation = gp,
  min_hap_samples = 3,
  show_reference_row = TRUE,
  gene_pos_x_angle = 90,
  text_size = 9
)
hap_figure
```

## Haplotype-phenotype association

`seed_weight` has a designed GeneA haplotype effect. `flowering_time` is a negative-control trait with the same distribution in all four designed groups.

```{r hap-pheno}
hap_pheno <- plot_hap_pheno(
  hap,
  phenotype = pheno,
  traits = c("seed_weight", "flowering_time"),
  min_hap_samples = 3,
  facet_ncol = 2,
  text_size = 9
)

hap_pheno$figure
hap_pheno$pvalue
```

## Single-variant phenotype association

`varA03` was designed so its ALT genotype is associated with higher `protein_content`.

```{r variant-pheno}
variant_pheno <- plot_variant_pheno(
  vcf,
  phenotype = pheno,
  variant_id = "varA03",
  traits = "protein_content",
  genotype_mode = "code",
  min_group_samples = 3,
  text_size = 9
)

variant_pheno$figure
variant_pheno$pvalue
```

## Linkage disequilibrium

`varLD01` through `varLD12` provide a deterministic LD-gradient example with a six-variant perfect core and progressively weaker downstream linkage.

```{r ld-block}
ld <- compute_ld_block(
  vcf,
  chrom = "chr1",
  start = 12342620,
  end = 12355500,
  variant_type = "snp",
  method = "r2",
  verbose = FALSE
)

ld$data
```

```{r ld-plot}
ld <- plot_ld_block(
  ld,
  annotation = gp,
  show_region = TRUE,
  show_variant_labels = FALSE,
  font = 9
)
ld$figure
```

The `GeneT` interval on chromosome 2 contains exactly two variants and is retained as the stable two-variant LD geometry case.

```{r ld-pair}
ld_pair <- compute_ld_block(
  vcf,
  chrom = "chr2",
  start = 16995001,
  end = 17006000,
  method = "r2",
  verbose = FALSE
)
ld_pair_figure <- plot_ld_block(ld_pair, return_object = FALSE, font = 9)
ld_pair_figure
```

## Phenotype-guided haplotype refinement

```{r refinement}
refined <- refine_haplotype(
  hap,
  phenotype = pheno,
  traits = "protein_content",
  min_hap_samples = 3,
  effect_threshold = 0.5
)

refined$refined_haplotypes
```

```{r refinement-plot}
refined_pheno <- plot_refined_hap_pheno(
  refined,
  phenotype = pheno,
  traits = "protein_content",
  min_hap_samples = 3,
  text_size = 9
)
refined_pheno$figure
refined_pheno$pvalue
```

The refined haplotype-variant plot returns a figure directly:

```{r refinement-variant-plot}
refined_variant_figure <- plot_refined_hap_variant(
  refined,
  annotation = gp,
  min_hap_samples = 3,
  text_size = 9
)
refined_variant_figure
```

## Variant-effect prioritization

```{r variant-effect}
effect_res <- plot_variant_effect(
  hap,
  phenotype = pheno,
  traits = "protein_content",
  min_group_samples = 3,
  x_axis = "position",
  text_size = 9
)

effect_res$figure
effect_res$effect
```

The deterministic model is intended to make documentation examples and regression tests share the same biological and statistical expectations rather than maintaining separate example datasets for each module.
