---
title: "Haplotype analysis"
author: "Rensc"
date: "2026-08-31"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{09 Haplotype analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
gtr_pkgdown <- tolower(Sys.getenv("GENETRACKR_PKGDOWN", "false")) %in%
  c("1", "true", "yes")
gtr_vignette_dpi <- if (gtr_pkgdown) 300 else 144

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12.5,
  fig.height = 8,
  dpi = gtr_vignette_dpi,
  fig.retina = 1,
  out.width = "100%",
  fig.align = "center"
)
```

## Introduction

```{r load-package}
library(GeneTrackR)
```

GeneTrackR builds haplotypes directly from VCF sample genotypes and keeps the complete analysis state in a `HapVariant` object. The stable workflow is to define the biological interval once, construct one canonical haplotype object, inspect the retained variants and samples, and pass that same object to phenotype association, refinement, and variant-effect analysis.

This article follows that object-centered workflow rather than rebuilding genotype matrices for each downstream task.

This article follows one continuous workflow:

1. prepare VCF genotype data and gene annotation;
2. build the primary gene-body haplotypes;
3. inspect the `HapVariant` object and its core tables;
4. choose an appropriate genotype representation;
5. control missing-genotype filtering;
6. construct transcript, flanking-region, or custom-region haplotypes;
7. draw the haplotype-variant figure;
8. pass the haplotype object to downstream phenotype and refinement workflows.

The main example uses the designed `GeneA` locus. Its gene body contains 11 variants and 36 complete samples that form four balanced haplotypes with nine samples per haplotype.

Package documentation: <https://renscq.github.io/GeneTrackR/>.

## Installation

After Bioconductor acceptance, install GeneTrackR with `BiocManager`:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("GeneTrackR")
library(GeneTrackR)
```

The development version remains available from the GeneTrackR GitHub repository.

## Step 1. Prepare VCF genotypes and annotation

Locate and read the bundled demo files:

```{r}
vcf_file <- system.file(
  "extdata",
  "gtr_demo_variants.vcf",
  package = "GeneTrackR"
)

gp_file <- system.file(
  "extdata",
  "gtr_demo.genePredExt",
  package = "GeneTrackR"
)

vcf <- read_vcf(
  vcf_file,
  mode = "memory",
  keep_genotype = TRUE,
  verbose = FALSE,
  progress = FALSE
)

gp <- read_genepred(
  gp_file,
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)
```

Haplotype construction requires VCF sample genotype columns. If a VCF was previously read with `keep_genotype = FALSE`, read it again with `keep_genotype = TRUE` before calling the haplotype functions.

For indexed large VCF files, the input can also be a lazy `VariantTrack` or a VCF file path. GeneTrackR retrieves only the requested haplotype region before constructing the genotype matrix.

## Step 2. Build the primary GeneA haplotypes

For gene- or transcript-defined analysis, use `hap_gene_variant()`. The primary GeneA workflow intentionally uses the **gene body only** so that the haplotype definition is based on variants inside the annotated locus:

```{r}
hap_gene <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  genotype_mode = "string"
)

hap_gene
```

The demo is designed so that this call contains:

- 11 GeneA gene-body variants;
- 36 retained samples;
- four haplotype groups;
- nine samples in each haplotype group.

The four genotype patterns are intentionally hierarchical: `Hap1/Hap2` are the closest pair, `Hap3/Hap4` are the second closest pair, and cross-pair comparisons differ at many more GeneA variants. This structure is used later by the phenotype-refinement demo.

Check these values directly from the returned object rather than assuming a fixed number for real datasets:

```{r}
hap_gene$meta
hap_gene$haplotypes[, c("hap_id", "sample_n", "samples"), with = FALSE]
```

`hap_id` is assigned from the observed genotype patterns after grouping. Treat the genotype pattern and sample membership as the biological result; do not assume that `Hap1`, `Hap2`, and so on have the same biological meaning across independently constructed regions.

## Step 3. Inspect the HapVariant object

A `HapVariant` keeps the complete state needed by downstream GeneTrackR functions:

```{r}
hap_gene$region
hap_gene$variants
hap_gene$genotype_long
hap_gene$genotype_wide
hap_gene$haplotypes
hap_gene$sample_haplotypes
```

The components have distinct purposes:

| Component | Content | Typical use |
| --- | --- | --- |
| `region` | locator, chromosome, boundaries, strand, and flank settings | confirm the biological interval |
| `variants` | retained VCF variant records | inspect variant order and REF/ALT alleles |
| `genotype_long` | one sample × variant genotype per row | detailed genotype inspection |
| `genotype_wide` | one sample per row with variant columns | inspect the full genotype matrix |
| `haplotypes` | one row per unique haplotype with sample counts | summarize haplotype groups |
| `sample_haplotypes` | sample-to-haplotype assignment | join with phenotype or other metadata |

The `meta` list records the genotype representation, missing-value policy, retained sample number, variant number, and haplotype number:

```{r}
hap_gene$meta
```

This object is already the canonical input for `plot_hap_variant()`, `plot_hap_pheno()`, `refine_haplotype()`, and `plot_variant_effect()`. There is no need to rebuild a genotype matrix manually.

## Step 4. Choose the genotype representation

GeneTrackR provides two genotype representations through `genotype_mode`.

### Allele-string mode

`genotype_mode = "string"` is recommended when the haplotype table itself is important for interpretation:

```{r}
hap_gene_string <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  genotype_mode = "string"
)

hap_gene_string$haplotypes
```

Single-base alleles are displayed directly as `A`, `T`, `C`, or `G`. Longer REF/ALT alleles are compressed to `iN`, where `N` is the displayed allele length. The haplotype plot maps allele classes in the stable order `A`, `T`, `C`, `G`, and `indel`.

String mode is also a compact display representation rather than a phased dosage representation: if a genotype contains an alternate allele, GeneTrackR displays the first observed ALT allele label for that variant. Inspect the original genotype columns in `vcf$data` when zygosity, phase, or multi-allelic dosage must be retained explicitly.

### Binary code mode

`genotype_mode = "code"` collapses each genotype to reference versus alternate presence:

```{r}
hap_gene_code <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  genotype_mode = "code"
)

hap_gene_code$haplotypes
```

In code mode:

- `0` means the genotype contains only the reference allele;
- `1` means at least one alternate allele is present.

Therefore code mode is a **binary REF/ALT-presence representation**, not allele dosage. A heterozygous genotype and a homozygous alternate genotype both become `1`. Use allele-string mode when the displayed allele identity is more important than compact binary grouping.

## Step 5. Control missing-genotype filtering

By default, `min_variant_number = NULL` requires a sample to have non-missing genotypes at **all retained variants**. This is the safest default for defining complete haplotypes.

The GeneA gene body has complete demo genotypes, so all 36 samples are retained:

```{r}
hap_gene$meta$sample_n
hap_gene$meta$min_variant_number
```

Flanking regions can introduce additional variants with missing data. The demo upstream variant `varAup01` is intentionally designed with missing/heterozygous genotypes. Adding it changes the haplotype definition:

```{r}
hap_gene_upstream <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  upstream = 500,
  downstream = 0,
  strand_aware = TRUE,
  genotype_mode = "string"
)

hap_gene_upstream$region
hap_gene_upstream$meta
```

To retain samples that are missing a limited number of variants, set an explicit threshold. The extended demo region contains 12 variants, so `min_variant_number = 11` allows one missing genotype:

```{r}
hap_gene_upstream_relaxed <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  upstream = 500,
  genotype_mode = "string",
  min_variant_number = 11
)

hap_gene_upstream_relaxed$sample_haplotypes
```

Use relaxed thresholds deliberately. Missing values become part of the observed haplotype pattern and may create additional haplotype groups. For the main GeneA tutorial, the unextended gene-body object `hap_gene` remains the recommended analysis object.

A custom missing label can be supplied for display:

```{r}
hap_missing_label <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  upstream = 500,
  genotype_mode = "code",
  missing_genotype = "-",
  min_variant_number = 11
)
```

Internally, GeneTrackR still tracks which genotypes are biologically missing, so a custom display label does not convert missing values into valid genotype calls.

## Step 6. Construct transcript, flanking-region, and custom-region haplotypes

### Transcript-defined haplotypes

Use `transcript_id` instead of `gene_id` when variants should be restricted to one transcript locus:

```{r}
hap_tx <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  transcript_id = "TxA1",
  genotype_mode = "string"
)

hap_tx$region
```

Supply exactly one of `gene_id` or `transcript_id`.

### Strand-aware flanking regions

`upstream` and `downstream` are interpreted relative to transcription direction when `strand_aware = TRUE`:

```{r}
hap_gene_flank <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  upstream = 1000,
  downstream = 500,
  strand_aware = TRUE,
  genotype_mode = "string",
  min_variant_number = 1
)

hap_gene_flank$region
```

For a negative-strand gene, biological upstream is toward increasing genomic coordinates. GeneTrackR handles this reversal automatically when `strand_aware = TRUE`.

### Explicit genomic regions

Use `hap_region_variant()` when the interval is defined directly rather than by annotation:

```{r}
hap_region <- hap_region_variant(
  vcf = vcf,
  chrom = "chr1",
  start = 12340001,
  end = 12352000,
  genotype_mode = "code"
)

hap_region
```

This is useful for GWAS/LD intervals, promoter windows, QTL regions, or any custom locus that does not correspond exactly to one annotated gene or transcript.

### Sample and variant-type filtering

The haplotype can be restricted before grouping. For example, analyze a subset of samples:

```{r}
hap_sample_subset <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  samples = sprintf("S%02d", 1:18),
  genotype_mode = "string"
)

hap_sample_subset$meta
```

Or construct haplotypes using only SNPs:

```{r}
hap_snp <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  variant_type = "SNP",
  genotype_mode = "string"
)

hap_snp$variants
```

Filtering variants changes the haplotype definition and may merge previously distinct haplotypes. Always inspect `haplotypes` and `sample_haplotypes` after changing the retained variant set.

`hap_variant()` remains available as a compatibility wrapper, but new code should prefer `hap_gene_variant()` for gene/transcript queries and `hap_region_variant()` for direct genomic intervals.

## Step 7. Draw the haplotype-variant figure

`plot_hap_variant()` combines the gene model, natural-variant markers, connector lines, REF/ALT reference rows, and haplotype genotype table:

```{r}
hap_variant_figure <- plot_hap_variant(
  hap_gene,
  annotation = gp,
  min_hap_samples = 3,
  show_reference_row = TRUE,
  variant_label = "pos",
  gene_pos_x_angle = 90,
  gene_track_legend_position = "top",
  direction_mode = "gene",
  table_x_angle = 90,
  text_size = 9
)

hap_variant_figure
```

General plotting palettes default to `Paired`; `plot_ld_block()` is the deliberate package-level exception and defaults to `Reds`. The main haplotype visual controls can still be changed independently:

```{r}
hap_variant_custom <- plot_hap_variant(
  hap_gene,
  annotation = gp,
  min_hap_samples = 3,
  show_reference_row = TRUE,
  gene_palette = "Paired",
  table_palette = "Paired",
  variant_palette = "Paired",
  table_alpha = 0.7,
  variant_alpha = 0.8,
  genotype_text_size = 3,
  variant_marker_size = 3,
  gene_track_height = 1.4,
  connector_height = 0.4,
  text_size = 9
)

hap_variant_custom
```

For allele-string haplotypes, table colors always follow the stable biological class order `A`, `T`, `C`, `G`, and `indel`. Custom colors can be supplied explicitly with those names:

```{r}
plot_hap_variant(
  hap_gene,
  annotation = gp,
  min_hap_samples = 3,
  table_colors = c(
    A = "#1B9E77",
    T = "#D95F02",
    C = "#7570B3",
    G = "#E7298A",
    indel = "#66A61E"
  ),
  text_size = 9
)
```

Use `show_variant_marker = FALSE` to hide natural-variant triangles, or adjust `variant_marker_size` when the locus contains many variants.

## Step 8. Pass the same haplotype object downstream

Do not reconstruct haplotypes separately for each downstream analysis. The `hap_gene` object created in Step 2 is the input for the following modules:

```{r}
# Haplotype-phenotype association
# hap_res <- plot_hap_pheno(
#   hap_gene,
#   phenotype = pheno,
#   traits = "seed_weight",
#   min_hap_samples = 3
# )

# Phenotype-guided refinement
# refined <- refine_haplotype(
#   hap_gene,
#   phenotype = pheno,
#   traits = "seed_weight",
#   min_hap_samples = 3
# )

# Variant-effect prioritization
# effect <- plot_variant_effect(
#   hap_gene,
#   phenotype = pheno,
#   traits = "protein_content",
#   min_group_samples = 3
# )
```

Keeping one well-defined `HapVariant` as the common input ensures that the variant set, sample set, missing-data rule, and haplotype assignment remain consistent across visualization, phenotype association, refinement, and variant-effect analysis.

## Session information

```{r session-info}
sessionInfo()
```
