---
title: "Strand-specific RNA-seq and Ribo-seq signal tracks"
author: "Rensc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Strand-specific RNA-seq and Ribo-seq signal tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5
)
```

This vignette uses the deterministic GeneTrackR demo genome. RNA-seq and Ribo-seq signals are each represented by separate plus- and minus-strand bedGraph files.

## Load annotation and signal files

```{r load-data}
library(GeneTrackR)

gp <- read_genepred(
  system.file("extdata", "gtr_demo.genePredExt", package = "GeneTrackR"),
  format = "genePredExt",
  verbose = FALSE
)

rnaseq_files <- system.file(
  "extdata",
  c("gtr_demo_rnaseq_plus.bedgraph", "gtr_demo_rnaseq_minus.bedgraph"),
  package = "GeneTrackR"
)

riboseq_files <- system.file(
  "extdata",
  c("gtr_demo_riboseq_plus.bedgraph", "gtr_demo_riboseq_minus.bedgraph"),
  package = "GeneTrackR"
)
```

bedGraph does not store strand metadata. The strand is therefore supplied explicitly when constructing the `BwgTrack` objects.

`read_bwg()` returns a `BwgTrack`. The signal plotting functions used below return ggplot/patchwork figures directly; they do not return a list with a `$figure` field. `merge_bwg()` returns another `BwgTrack`, so the merged object can be passed directly into any signal plotting function.

```{r read-signal}
rnaseq <- read_bwg(
  rnaseq_files,
  format = "bedgraph",
  sample_names = c("RNA_seq_plus", "RNA_seq_minus"),
  strand = c("+", "-"),
  mode = "memory",
  verbose = FALSE
)

riboseq <- read_bwg(
  riboseq_files,
  format = "bedgraph",
  sample_names = c("Ribo_seq_plus", "Ribo_seq_minus"),
  strand = c("+", "-"),
  mode = "memory",
  verbose = FALSE
)

rnaseq$samples[, c("sample_id", "strand"), with = FALSE]
riboseq$samples[, c("sample_id", "strand"), with = FALSE]
```

## RNA-seq: exon-enriched coverage

The RNA-seq demo tracks contain signal over exons, including UTRs. Designed intronic and intergenic regions have no bedGraph records. `GeneA` is on the plus strand and `GeneB` is on the minus strand.

```{r rnaseq-genea}
rnaseq_genea_figure <- plot_signal_gene(
  signal = rnaseq,
  annotation = gp,
  gene_id = "GeneA",
  plot_type = "bar",
  signal_y_scale = "fixed"
)
rnaseq_genea_figure
```

```{r rnaseq-geneb}
rnaseq_geneb_figure <- plot_signal_gene(
  signal = rnaseq,
  annotation = gp,
  gene_id = "GeneB",
  plot_type = "bar",
  signal_y_scale = "fixed"
)
rnaseq_geneb_figure
```


## Palette direction and panel proportions

`signal_palette_direction` reverses the standard palette order before discrete colors are assigned. The effect is visible even when strand-aware retrieval leaves a single signal sample. Explicit `signal_colors` remain unchanged because they are treated as exact user mappings.

```{r rnaseq-palette-layout}
rnaseq_palette_figure <- plot_signal_gene(
  signal = rnaseq,
  annotation = gp,
  gene_id = "GeneA",
  plot_type = "bar",
  signal_palette = "Blues",
  signal_palette_direction = -1,
  signal_y_scale = "fixed",
  signal_track_height = 4,
  gene_track_height = 1
)
rnaseq_palette_figure
```

The same height arguments are available in `plot_signal_transcript()` and `plot_signal_region()`. The integrated `plot_tracks()` function already exposes a named `heights` vector.

## Ribo-seq: sparse base-resolution P-site counts

The Ribo-seq tracks contain moderately dense one-base integer P-site-like counts only in CDS regions. Frame 0 is broadly occupied with variable heights, while frame 1 and frame 2 use different codon subsets with lower irregular counts. Zero-count bases are omitted. Counts follow transcript orientation, are distributed at approximately 80%/10%/10% across frame 0/frame 1/frame 2, and initiation/termination frame-0 counts are approximately two times the internal frame-0 mean.

For the positive-strand `TxA1` transcript:

```{r riboseq-plus}
riboseq_plus_figure <- plot_signal_transcript(
  signal = riboseq,
  annotation = gp,
  transcript_id = "TxA1",
  coordinate = "transcript",
  plot_type = "frame"
)
riboseq_plus_figure
```

For the negative-strand `TxB1` transcript:

```{r riboseq-minus}
riboseq_minus_figure <- plot_signal_transcript(
  signal = riboseq,
  annotation = gp,
  transcript_id = "TxB1",
  coordinate = "transcript",
  plot_type = "frame"
)
riboseq_minus_figure
```

## Combine all four signal tracks

RNA-seq and Ribo-seq objects have distinct sample IDs and can therefore be merged directly.

```{r merge-tracks}
signal_all <- merge_bwg(rnaseq, riboseq)
summary_bwg(signal_all)
```

A regional plot can display both assays and both strands over the GeneA-GeneB-GeneC locus.

```{r regional-plot}
regional_signal_figure <- plot_signal_region(
  signal = signal_all,
  annotation = gp,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  strand = "both",
  plot_type = "bar"
)
regional_signal_figure
```

The same `BwgTrack` can be passed directly to `plot_tracks()`. Here it is combined with the GeneA annotation track:

```{r integrated-tracks}
integrated_signal_figure <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  gene_id = "GeneA",
  signal_type = "bar"
)
integrated_signal_figure
```
