---
title: "Variant tracks"
description: "Read, validate, query, filter, write, and visualize VCF-derived variants with VariantTrack objects."
author: "Rensc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variant tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.8,
  dpi = 500
)
```

GeneTrackR uses `VariantTrack` as the common object for VCF-derived genomic variants. A normal variant workflow is:

1. read a VCF into memory or create a lazy indexed track;
2. validate the standardized variant records;
3. summarize the current collection;
4. retrieve variants by genomic region, gene, transcript, ID, type, or text pattern;
5. visualize the selected variants;
6. pass the resulting `VariantTrack` to haplotype, LD, phenotype, or integrated browser workflows.

The examples below use the same `GeneA` / `TxA1` locus used throughout the GeneTrackR documentation.

```{r load-package}
library(GeneTrackR)

vcf_file <- system.file(
  "extdata",
  "gtr_demo_variants.vcf",
  package = "GeneTrackR"
)

gp <- read_genepred(
  system.file(
    "extdata",
    "gtr_demo.genePredExt",
    package = "GeneTrackR"
  ),
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)
```

## Read variants

For a small VCF or repeated local analysis, use `mode = "memory"` to parse all records immediately.

```{r read-vcf}
variants <- read_vcf(
  vcf_file,
  mode = "memory",
  keep_genotype = TRUE,
  verbose = FALSE,
  progress = FALSE
)

variants
```

`read_vcf()` returns a `VariantTrack`. Standard VCF fields are normalized to columns including `chrom`, `pos`, `variant_id`, `ref`, `alt`, `qual`, `filter`, `info`, and `variant_type`. Genotype fields and sample columns are retained when `keep_genotype = TRUE`.

```{r inspect-variant-track}
head(variants$data[, c(
  "chrom",
  "pos",
  "variant_id",
  "ref",
  "alt",
  "variant_type"
)])

variants$meta$sample_names
```

`VariantTrack` uses **1-based genomic positions**, matching the VCF POS convention.

### Large indexed VCF files

For a bgzip-compressed VCF with a `.tbi` or `.csi` index, `mode = "auto"` or `mode = "lazy"` can avoid parsing the complete file. The lazy object stores source metadata first and reads records only when a genomic interval is requested.

```{r lazy-vcf, eval=FALSE}
large_vcf <- "/path/to/large.vcf.gz"

variants_lazy <- read_vcf(
  large_vcf,
  mode = "lazy"
)

variants_region <- retrieve_vcf(
  variants_lazy,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  as = "VariantTrack"
)
```

Lazy regional access requires an indexed VCF and `Rsamtools`. If `mode = "lazy"` is requested for a non-indexed VCF, GeneTrackR falls back to memory mode with a warning.

## Validate and summarize

Use `validate_vcf()` before downstream analysis when VCF inputs come from heterogeneous pipelines or have been modified outside GeneTrackR.

```{r validate-vcf}
validation <- validate_vcf(variants)

validation$invalid_summary
validation$warnings
```

The validation report separates record-level failures from general warnings. It checks required identifiers and allele fields, genomic positions, duplicated variant IDs, and consistency among `pos`, `start`, and `end` when those fields are available.

`summary_vcf()` reports counts for the current variant collection. The default grouping is chromosome plus inferred variant type.

```{r summarize-vcf}
summary_vcf(variants)
summary_vcf(variants, chrom = "chr1")
```

Custom grouping is possible when the requested fields exist in the standardized table.

```{r summarize-filter}
summary_vcf(
  variants,
  by = c("filter", "variant_type")
)
```

## Retrieve by genomic region

`retrieve_vcf()` returns either a plain `data.table` for direct inspection or another `VariantTrack` for downstream GeneTrackR operations.

```{r retrieve-region-table}
region_table <- retrieve_vcf(
  variants,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  as = "data.table",
  verbose = FALSE
)

region_table[, .(
  chrom,
  pos,
  variant_id,
  ref,
  alt,
  variant_type
)]
```

Use `as = "VariantTrack"` when the subset will be plotted or passed into haplotype, phenotype, or LD functions.

```{r retrieve-region-track}
region_variants <- retrieve_vcf(
  variants,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  as = "VariantTrack",
  verbose = FALSE
)

region_variants
```

A direct region query must use one chromosome with both `start` and `end`. A chromosome-only query is also valid when coordinates are not required.

## Retrieve by gene or transcript

When annotation is available, `retrieve_vcf()` can resolve a gene or transcript into its genomic interval.

```{r retrieve-gene}
genea_variants <- retrieve_vcf(
  variants,
  annotation = gp,
  gene_id = "GeneA",
  as = "VariantTrack",
  verbose = FALSE
)

genea_variants
```

The deterministic demo contains 11 variants within the `GeneA` gene body. Flanking sequence can be added with `upstream` and `downstream`.

```{r retrieve-gene-flank}
genea_extended <- retrieve_vcf(
  variants,
  annotation = gp,
  gene_id = "GeneA",
  upstream = 1000,
  downstream = 500,
  strand_aware = TRUE,
  as = "VariantTrack",
  verbose = FALSE
)

genea_extended
```

With `strand_aware = TRUE`, upstream and downstream follow transcriptional direction. This distinction matters for negative-strand genes.

Transcript-aware retrieval uses the same interface.

```{r retrieve-transcript}
txa1_variants <- retrieve_vcf(
  variants,
  annotation = gp,
  transcript_id = "TxA1",
  as = "VariantTrack",
  verbose = FALSE
)

txa1_variants
```

Do not mix `gene_id` or `transcript_id` with direct `chrom` / `start` / `end` arguments in one query.

## Filter by variant attributes

Variant ID, inferred variant type, and text pattern filters can be used independently or combined with a genomic locator.

```{r filter-ids}
selected_variants <- retrieve_vcf(
  variants,
  variant_id = c("varA03", "varA04", "varA05"),
  as = "data.table",
  verbose = FALSE
)

selected_variants[, .(
  chrom,
  pos,
  variant_id,
  ref,
  alt,
  variant_type
)]
```

```{r filter-type}
genea_indels <- retrieve_vcf(
  genea_variants,
  variant_type = c("INS", "DEL"),
  as = "data.table",
  verbose = FALSE
)

genea_indels[, .(
  pos,
  variant_id,
  ref,
  alt,
  variant_type
)]
```

`pattern` searches the standardized variant ID, REF, ALT, INFO, and variant-type fields. Use `fixed = TRUE` when the search term should not be interpreted as a regular expression.

```{r filter-pattern}
high_ld_variants <- retrieve_vcf(
  variants,
  pattern = "high_ld",
  fixed = TRUE,
  as = "VariantTrack",
  verbose = FALSE
)

high_ld_variants
```

## Plot a variant track

`plot_variant()` is the dedicated variant plotting function. It returns a ggplot object and supports lollipop, point, and rug geometries.

```{r plot-gene-variants}
variant_figure <- plot_variant(
  genea_variants,
  color_by = "variant_type",
  label_by = "variant_id",
  variant_shape = "lollipop",
  variant_palette = "Paired",
  point_size = 2.5,
  text_size = 12,
  plot_theme = "classic"
)

variant_figure
```

The primary controls are:

| Argument | Purpose |
|---|---|
| `color_by` | Color by `variant_type`, `filter`, or disable grouping. |
| `label_by` | Label by variant ID, type, REF, ALT, or hide labels. |
| `variant_shape` | Draw `lollipop`, `point`, or `rug` geometry. |
| `variant_palette` | Select an RColorBrewer palette. |
| `variant_colors` | Supply explicit named or positional colors. |

For stable biological classes across multiple figures, prefer explicit named colors.

```{r plot-explicit-colors}
plot_variant(
  genea_variants,
  color_by = "variant_type",
  label_by = "none",
  variant_shape = "rug",
  variant_colors = c(
    SNP = "#1F78B4",
    INS = "#33A02C",
    DEL = "#E31A1C"
  ),
  plot_theme = "classic"
)
```

`plot_variant_track()` is retained as a backward-compatible wrapper around `plot_variant()`.

## Write a standardized VCF subset

`write_vcf()` writes the standardized site-level VCF fields from a `VariantTrack`.

```{r write-vcf, eval=FALSE}
write_vcf(
  genea_variants,
  "GeneA.variants.vcf",
  overwrite = TRUE
)
```

The current writer exports CHROM, POS, ID, REF, ALT, QUAL, FILTER, and INFO. It does **not** round-trip FORMAT and genotype sample columns. Keep the original genotype-containing VCF as the archival source whenever genotype preservation is required.

## Downstream handoff

The same `VariantTrack` object can be reused directly by later GeneTrackR modules:

- `hap_gene_variant()` or `hap_region_variant()` for haplotype construction;
- `plot_variant_pheno()` for single-variant phenotype association;
- `compute_ld_block()` and `plot_ld_block()` for LD analysis;
- `plot_tracks()` for integrated browser-style visualization.

Keep genotype columns by reading with `keep_genotype = TRUE` whenever downstream haplotype or LD analysis is planned.
