## Installation

GeneTrackR depends on CRAN and Bioconductor packages. Install the dependencies first, then install GeneTrackR from GitHub. GeneTrackR is a pure-R package and does not require a C/C++ compilation toolchain.

```{r}
## CRAN dependencies
install.packages(c(
  "devtools",
  "data.table",
  "ggplot2",
  "patchwork",
  "rlang",
  "RColorBrewer"
))

## Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c(
  "GenomicRanges",
  "IRanges",
  "S4Vectors"
), ask = FALSE, update = FALSE)

## Optional indexed-query acceleration
BiocManager::install("Rsamtools", ask = FALSE, update = FALSE)

## Install GeneTrackR from GitHub
devtools::install_github(
  "Renscq/GeneTrackR",
  dependencies = TRUE,
  build_vignettes = FALSE
)
```

`Rsamtools` is optional and is only required when GeneTrackR uses the Bioconductor tabix backend for indexed VCF or bedGraph regional queries.

Load the package:

```{r}
library(GeneTrackR)
```

For local development:

```bash
git clone https://github.com/Renscq/GeneTrackR.git
cd GeneTrackR
```

```{r}
devtools::document()
devtools::install()
```
