## Installation

GeneTrackR 0.99.2 is the current Bioconductor submission candidate. After acceptance, the recommended installation path is `BiocManager::install("GeneTrackR")`. GeneTrackR is a pure-R package and does not require a GeneTrackR-specific C/C++ compilation toolchain.

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

## Available after Bioconductor acceptance
BiocManager::install("GeneTrackR")
```

Until the package is accepted by Bioconductor, install the development source from GitHub. `BiocManager` resolves the Bioconductor dependencies while `remotes` installs the source repository.

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

BiocManager::install(c(
  "GenomicRanges",
  "IRanges",
  "S4Vectors"
), ask = FALSE, update = FALSE)

remotes::install_github(
  "Renscq/GeneTrackR",
  dependencies = TRUE,
  build_vignettes = FALSE
)
```

`Rsamtools` is an optional dependency used only for indexed VCF or bedGraph regional queries. Install it when tabix acceleration is required:

```{r}
BiocManager::install("Rsamtools", ask = FALSE, update = FALSE)
```

Load the package:

```{r}
library(GeneTrackR)
```

For local development:

```bash
git clone https://github.com/Renscq/GeneTrackR.git
cd GeneTrackR
```

```{r}
devtools::document()
devtools::install()
```
