## Annotation files: read, subset, merge, and write

GeneTrackR supports five annotation inputs: **GenePred**, **GenePredExt**, **GTF**, **GFF3**, and **BED**. All readers standardize coordinates internally to **1-based closed** coordinates, but the source formats are not equivalent: GenePred/GenePredExt, GTF, and GFF3 can provide hierarchical gene models, whereas BED is primarily treated as an interval annotation track.

### Supported formats and object types

| Format | Typical content | Reader | Main return | Writer format |
|---|---|---|---|---|
| GenePred | 10-column transcript model | `read_genepred(..., format = "genePred")` | `GenePred` | `genepred` |
| GenePredExt | 15-column GenePred with gene ID/status/frame fields | `read_genepred(..., format = "genePredExt")` | `GenePred` | `genepredext` |
| GTF | 9-column hierarchical annotation with attributes | `read_gtf()` | `FeatureTrack` with derived gene model | `gtf` |
| GFF3 | 9-column hierarchical annotation with `ID`/`Parent` relationships | `read_gff()` | `FeatureTrack` with derived gene model | `gff` |
| BED | BED3-BED12 interval annotation | `read_bed()` | interval-style `FeatureTrack` | `bed6`; `bed12` for gene-model objects |

Coordinate conventions are handled at the file boundary:

- GenePred/GenePredExt default to UCSC-style **0-based half-open** input with `coordinate = "ucsc"`.
- BED defaults to **0-based half-open** input with `coordinate = "bed"`.
- GTF and GFF3 use **1-based closed** coordinates.
- GeneTrackR stores all annotation coordinates internally as **1-based closed** coordinates.
- `write_feature()` converts GenePred/BED-like outputs back to UCSC coordinates by default with `coordinate = "ucsc"`.

The built-in annotation files describe the same demo genome:

```{r}
gp_file <- system.file(
  "extdata",
  "gtr_demo.genePredExt",
  package = "GeneTrackR"
)
gtf_file <- system.file("extdata", "gtr_demo.gtf", package = "GeneTrackR")
gff_file <- system.file("extdata", "gtr_demo.gff3", package = "GeneTrackR")
bed_file <- system.file(
  "extdata",
  "gtr_demo_features.bed",
  package = "GeneTrackR"
)
```

### GenePred and GenePredExt

GenePred stores one transcript per row with transcript/CDS boundaries and exon blocks. Standard GenePred contains 10 core columns. GenePredExt adds `score`, `name2`, CDS status, and exon-frame information; GeneTrackR normally uses `name` as `transcript_id` and `name2` as `gene_id` for GenePredExt.

#### Read

The built-in file is GenePredExt:

```{r}
gp <- read_genepred(
  gp_file,
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)

gp
head(gp$genes)
head(gp$transcripts)
head(gp$exons)
```

Standard GenePred uses the same reader. Here a standard GenePred file is created from the demo object and read back:

```{r}
gp_standard_file <- file.path(tempdir(), "gtr_demo.genePred")

write_feature(
  gp,
  gp_standard_file,
  format = "genepred",
  overwrite = TRUE
)

gp_standard <- read_genepred(
  gp_standard_file,
  format = "genePred",
  verbose = FALSE,
  progress = FALSE
)
```

#### Subset

GenePred-compatible objects can be subset by gene, transcript, or genomic region:

```{r}
gp_gene_a <- retrieve_feature(
  gp,
  gene_id = "GeneA"
)

gp_tx_a1 <- retrieve_feature(
  gp,
  transcript_id = "TxA1"
)

gp_region <- retrieve_feature(
  gp,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  mode = "overlap"
)
```

For a plain table instead of a sub-object:

```{r}
gp_gene_table <- retrieve_feature(
  gp,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  level = "gene",
  as = "data.table"
)
```

#### Merge

Subsets from the same or different GenePred objects can be recombined with `merge_feature()`:

```{r}
gp_gene_b <- retrieve_feature(gp, gene_id = "GeneB")

gp_merged <- merge_feature(
  gp_gene_a,
  gp_gene_b,
  source_names = c("GeneA", "GeneB"),
  conflict = "deduplicate"
)
```

#### Write

A GenePred-compatible object can be written as either GenePred or GenePredExt:

```{r}
write_feature(
  gp_gene_a,
  file.path(tempdir(), "GeneA.genePred"),
  format = "genepred",
  overwrite = TRUE
)

write_feature(
  gp_gene_a,
  file.path(tempdir(), "GeneA.genePredExt"),
  format = "genepredext",
  overwrite = TRUE
)
```

`write_genepred()` remains available as a backward-compatible wrapper around `write_feature()`.

### GTF

GTF stores gene-model records in a 9-column table and uses the final attribute column for identifiers such as `gene_id` and `transcript_id`. GeneTrackR parses the hierarchical records into a `FeatureTrack` with standardized `$genes`, `$transcripts`, `$exons`, and `$data` tables when the required records are present.

#### Read

```{r}
gtf <- read_gtf(
  gtf_file,
  verbose = FALSE,
  progress = FALSE
)

gtf
head(gtf$genes)
head(gtf$transcripts)
```

Selected feature types can be loaded when the full annotation is not required:

```{r}
gtf_core <- read_gtf(
  gtf_file,
  feature_types = c("gene", "transcript", "exon", "CDS"),
  verbose = FALSE,
  progress = FALSE
)
```

#### Subset

```{r}
gtf_gene_a <- retrieve_feature(gtf, gene_id = "GeneA")

gtf_gene_a_exons <- retrieve_feature(
  gtf,
  gene_id = "GeneA",
  level = "exon",
  as = "data.table"
)

gtf_region <- retrieve_feature(
  gtf,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  mode = "overlap"
)
```

#### Merge

```{r}
gtf_gene_b <- retrieve_feature(gtf, gene_id = "GeneB")

gtf_merged <- merge_feature(
  gtf_gene_a,
  gtf_gene_b,
  source_names = c("GeneA", "GeneB")
)
```

#### Write

```{r}
write_feature(
  gtf_gene_a,
  file.path(tempdir(), "GeneA.gtf"),
  format = "gtf",
  overwrite = TRUE
)
```

Because a gene-model GTF contains transcript and exon hierarchy, it can also be converted directly to GenePred/GenePredExt or BED12 with `write_feature()`.

### GFF3

GFF3 also uses 9 columns, but hierarchy is represented primarily through `ID` and `Parent` attributes. `read_gff()` reconstructs gene/transcript/exon relationships and returns the same unified `FeatureTrack` interface used for GTF.

#### Read

```{r}
gff <- read_gff(
  gff_file,
  verbose = FALSE,
  progress = FALSE
)

gff
head(gff$genes)
head(gff$transcripts)
```

Feature-type filtering is also supported:

```{r}
gff_core <- read_gff(
  gff_file,
  feature_types = c("gene", "mRNA", "exon", "CDS"),
  verbose = FALSE,
  progress = FALSE
)
```

#### Subset

```{r}
gff_gene_a <- retrieve_feature(gff, gene_id = "GeneA")

gff_region <- retrieve_feature(
  gff,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  mode = "overlap"
)
```

#### Merge

```{r}
gff_gene_b <- retrieve_feature(gff, gene_id = "GeneB")

gff_merged <- merge_feature(
  gff_gene_a,
  gff_gene_b,
  source_names = c("GeneA", "GeneB")
)
```

#### Write

```{r}
write_feature(
  gff_gene_a,
  file.path(tempdir(), "GeneA.gff3"),
  format = "gff",
  overwrite = TRUE
)
```

### BED

BED is treated differently from GenePred/GTF/GFF3. `read_bed()` reads BED3-BED12-style files as **interval annotations** and standardizes the interval coordinates, name, score, and strand fields into a `FeatureTrack`. It does not reconstruct BED12 block columns into a transcript/exon hierarchy; use GenePred, GTF, or GFF3 when exon-level gene models are required.

#### Read

```{r}
features <- read_bed(
  bed_file,
  verbose = FALSE,
  progress = FALSE
)

features
head(features$data)
```

The demo BED contains promoters, enhancers, candidate regions, repeats, conserved regions, and QTL intervals. Their semantic type is encoded in the BED name field, for example `GeneA_promoter|promoter`.

#### Subset

BED intervals can be selected by region or text pattern:

```{r}
bed_gene_a <- retrieve_feature(
  features,
  chrom = "chr1",
  start = 12338201,
  end = 12352000,
  mode = "overlap"
)

bed_promoters <- retrieve_feature(
  features,
  pattern = "promoter"
)
```

#### Merge

```{r}
bed_gene_b <- retrieve_feature(
  features,
  chrom = "chr1",
  start = 12355501,
  end = 12368500,
  mode = "overlap"
)

bed_merged <- merge_feature(
  bed_gene_a,
  bed_gene_b,
  source_names = c("GeneA_region", "GeneB_region"),
  conflict = "keep_all"
)
```

#### Write

Generic BED-derived intervals should normally be written as BED6:

```{r}
write_feature(
  bed_merged,
  file.path(tempdir(), "merged_features.bed6"),
  format = "bed6",
  overwrite = TRUE
)
```

BED12 output requires transcript/exon gene-model information. It is therefore appropriate for `GenePred` or gene-model GTF/GFF3 objects rather than a generic BED interval track:

```{r}
write_feature(
  gp_gene_a,
  file.path(tempdir(), "GeneA.bed12"),
  format = "bed12",
  overwrite = TRUE
)
```

### Region-selection modes

`retrieve_feature()` uses the same region-selection interface across annotation formats:

| `mode` | Meaning |
|---|---|
| `"overlap"` | Keep records overlapping the requested region. This is the default. |
| `"within"` | Keep records fully contained in the requested region. |
| `"trim"` | For hierarchical gene models, keep overlapping records and clip transcript/exon boundaries to the requested interval. |

For generic BED-style interval tracks, `"trim"` behaves as overlap selection because there is no transcript/exon hierarchy to rebuild.

### Merge conflict handling

`merge_feature()` accepts `Feature`, `FeatureTrack`, and `GenePred` objects, including a list of objects. Duplicate identifiers are handled explicitly:

| `conflict` | Behavior |
|---|---|
| `"deduplicate"` | Keep the first compatible copy according to input order. Default. |
| `"rename"` | Rename conflicting IDs in later inputs and update hierarchy references. |
| `"keep_all"` | Retain conflicting records unchanged. |
| `"error"` | Stop when duplicate IDs are detected. |
| `"keep_first"` | Backward-compatible alias of `"deduplicate"`. |

When the same locus is loaded from different annotation formats, `rename` is useful if both representations should be retained:

```{r}
cross_format <- merge_feature(
  list(gp_gene_a, gtf_gene_a),
  source_names = c("GenePred", "GTF"),
  conflict = "rename"
)
```

If the two inputs represent the same annotation and only one copy is needed, use `conflict = "deduplicate"` instead.

### Cross-format conversion and standardized tables

The readers share a common internal representation, so annotation formats can be converted without reparsing text manually:

```{r}
# GTF/GFF3 gene models -> GenePred-compatible object
gtf_as_gp <- as_genepred(gtf)
gff_as_gp <- as_genepred(gff)

# Any annotation object -> standardized feature table
feature_dt <- as_feature_table(gp)

# Gene-model tables
gene_dt <- as_gene_table(gp)
tx_dt <- as_transcript_table(gp)
exon_dt <- as_exon_table(gp)

# Bioconductor interoperability
gene_gr <- as_granges(gp, level = "gene")
```

`write_feature()` is also the main cross-format writer. For example, a GTF-derived gene model can be exported as GenePredExt or BED12:

```{r}
write_feature(
  gtf_gene_a,
  file.path(tempdir(), "GeneA.from_gtf.genePredExt"),
  format = "genepredext",
  overwrite = TRUE
)

write_feature(
  gtf_gene_a,
  file.path(tempdir(), "GeneA.from_gtf.bed12"),
  format = "bed12",
  overwrite = TRUE
)
```

### Annotation object contract

After reading, GeneTrackR keeps annotation data in a small set of compatible objects:

| Input | Main object | Important contents |
|---|---|---|
| GenePred / GenePredExt | `GenePred` | `$genes`, `$transcripts`, `$exons`, `$data` |
| Gene-model GTF / GFF3 | `FeatureTrack` | `$data` plus derived `$genes`, `$transcripts`, `$exons` |
| BED | interval `FeatureTrack` | primarily `$data` |

`retrieve_feature()` returns a compatible sub-object by default (`as = "Feature"`) and a `data.table` when `as = "data.table"`. `merge_feature()` always returns a unified `Feature` object and additionally inherits from `GenePred` when transcript/exon tables are available.

### How annotation objects connect to the rest of GeneTrackR

Annotation objects are the first input to most downstream workflows. `GenePred` or gene-model `FeatureTrack` objects can be passed directly to gene-model plotting, signal plotting, haplotype extraction, LD visualization, and integrated browser-style tracks.

### Core object flow and return contracts

GeneTrackR uses a small set of S3 objects throughout the workflow. The examples below keep these objects intact instead of repeatedly converting them to plain tables.

| Step | Function | Main return | Important contents / next consumer |
|---|---|---|---|
| Annotation | `read_genepred()` | `GenePred` | `$genes`, `$transcripts`, `$exons`, `$data`; used by gene/signal/haplotype plotting |
| GTF/GFF/BED | `read_gtf()`, `read_gff()`, `read_bed()` | `FeatureTrack` | `$data` plus derived hierarchy when available; gene-model FeatureTracks are GenePred-convertible |
| Signal | `read_bwg()` / `merge_bwg()` | `BwgTrack` | `$samples`, optional in-memory `$data`, `$meta`; consumed by `plot_signal_*()` and `plot_tracks()` |
| Variant | `read_vcf()` | `VariantTrack` | `$data`, `$meta`; consumed by haplotype, LD, phenotype, and track functions |
| Variant subset | `retrieve_vcf()` | `data.table` by default; `VariantTrack` with `as = "VariantTrack"` | use the table for inspection or request `VariantTrack` for downstream track/object workflows |
| Haplotype | `hap_gene_variant()` / `hap_region_variant()` | `HapVariant` | `$variants`, `$haplotypes`, `$sample_haplotypes`, `$genotype_wide` |
| Phenotype association | `plot_hap_pheno()` / `plot_variant_pheno()` | `GeneTrackRPhenoPlot` | `$figure`, `$pvalue`, `$summary`, `$plot_data` |
| LD | `compute_ld_block()` | `LDTrack` | `$data`, `$matrix`, `$variants`, `$region`; `plot_ld_block()` stores the figure in `$figure` |
| Refinement | `refine_haplotype()` | `HapRefined` | `$refined_hap`, `$refined_haplotypes`, `$haplotype_map`, `$pairwise_test` |
| Variant effect | `plot_variant_effect()` | `GeneTrackRVariantEffectPlot` | `$figure`, `$effect`, `$plot_data` |

Plotting functions such as `plot_gene()`, `plot_signal_gene()`, `plot_variant()`, `plot_tracks()`, `plot_hap_variant()`, and `plot_refined_hap_variant()` return a ggplot/patchwork figure directly. In contrast, phenotype/effect plotting functions return result objects containing `$figure` plus analysis tables. `plot_ld_block()` is intentionally stateful: by default it returns an updated `LDTrack` with the plot stored in `$figure`; use `return_object = FALSE` only when a figure-only return is required.
