## RNA-seq and Ribo-seq signal tracks

GeneTrackR uses `BwgTrack` objects for continuous genomic signals. This module follows one complete workflow for the strand-specific RNA-seq and Ribo-seq demo data:

1. read bedGraph files;
2. write bigWig files;
3. plot gene-level signal tracks;
4. plot transcript-level signal tracks;
5. combine RNA-seq and Ribo-seq with `plot_tracks()`;
6. plot signals across a genomic region.

The examples use the same `GeneA`/`TxA1` positive-strand locus whenever possible so that the outputs can be compared directly between steps.

### Step 1. Read RNA-seq and Ribo-seq bedGraph data

Load the annotation and locate the four strand-specific signal files:

```{r}
gp_file <- system.file(
  "extdata",
  "gtr_demo.genePredExt",
  package = "GeneTrackR"
)

rnaseq_files <- system.file(
  "extdata",
  c(
    "gtr_demo_rnaseq_plus.bedgraph",
    "gtr_demo_rnaseq_minus.bedgraph"
  ),
  package = "GeneTrackR"
)

riboseq_files <- system.file(
  "extdata",
  c(
    "gtr_demo_riboseq_plus.bedgraph",
    "gtr_demo_riboseq_minus.bedgraph"
  ),
  package = "GeneTrackR"
)

gp <- read_genepred(
  gp_file,
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)
```

Read the RNA-seq plus/minus bedGraph files into one `BwgTrack` object:

```{r}
rnaseq <- read_bwg(
  rnaseq_files,
  format = "bedgraph",
  sample_names = c("RNA_seq_plus", "RNA_seq_minus"),
  strand = c("+", "-"),
  mode = "memory"
)

rnaseq
summary_bwg(rnaseq)
```

Read the Ribo-seq plus/minus bedGraph files in the same way:

```{r}
riboseq <- read_bwg(
  riboseq_files,
  format = "bedgraph",
  sample_names = c("Ribo_seq_plus", "Ribo_seq_minus"),
  strand = c("+", "-"),
  mode = "memory"
)

riboseq
summary_bwg(riboseq)
```

The two assays intentionally have different signal structures:

- **RNA-seq** signal is enriched across exons, including UTRs, while intronic and intergenic regions have little or no coverage in the demo data.
- **Ribo-seq** signal uses moderately dense 1-bp integer P-site-like counts within CDS regions. Frame 0 is broadly occupied with variable heights; frame 1 and frame 2 occur at different subsets of codons with lower irregular counts. Zero-count bases are omitted from bedGraph. Total counts remain approximately 80%/10%/10% for frame 0/frame 1/frame 2, and the initiation/termination frame-0 counts are approximately two times the internal frame-0 mean.

Because bedGraph does not store strand metadata, `strand = c("+", "-")` explicitly records the strand associated with each input file in the `BwgTrack` sample table.

### Step 2. Write RNA-seq and Ribo-seq bigWig files

`write_bwg()` can convert an in-memory `BwgTrack` to bigWig directly with the bundled third-party libBigWig library in `src/`. No external conversion program is required; only chromosome sizes are needed for the bigWig header.

```{r}
chrom_sizes_file <- system.file(
  "extdata",
  "gtr_demo.chrom.sizes",
  package = "GeneTrackR"
)

bigwig_dir <- file.path(tempdir(), "GeneTrackR_demo_bigwig")
dir.create(bigwig_dir, recursive = TRUE, showWarnings = FALSE)
```

Write the RNA-seq tracks:

```{r}
rnaseq_bigwig <- write_bwg(
  rnaseq,
  outdir = bigwig_dir,
  format = "bigwig",
  chrom_sizes = chrom_sizes_file,
  overwrite = TRUE
)

rnaseq_bigwig
```

Write the Ribo-seq tracks:

```{r}
riboseq_bigwig <- write_bwg(
  riboseq,
  outdir = bigwig_dir,
  format = "bigwig",
  chrom_sizes = chrom_sizes_file,
  overwrite = TRUE
)

riboseq_bigwig
```

Each call invisibly returns a table with `sample_id`, output `file`, and `format`. With the sample names used above, the output directory contains:

```text
RNA_seq_plus.bigwig
RNA_seq_minus.bigwig
Ribo_seq_plus.bigwig
Ribo_seq_minus.bigwig
```

BigWig export has a single backend in GeneTrackR: the bundled libBigWig implementation. This keeps the write path deterministic across platforms and avoids an external executable dependency.

### Step 3. Plot RNA-seq and Ribo-seq tracks for a gene

`plot_signal_gene()` retrieves the genomic span of a gene and optionally adds the gene model below the signal panel. With `strand = "auto"`, the gene strand is used to select the matching signal track. `GeneA` is on the positive strand, so the following examples use the plus RNA-seq and Ribo-seq samples automatically.

RNA-seq gene track:

```{r}
p_rnaseq_gene <- plot_signal_gene(
  signal = rnaseq,
  annotation = gp,
  gene_id = "GeneA",
  plot_type = "bar",
  strand = "auto",
  signal_palette = "Blues",
  signal_palette_direction = -1,
  signal_y_scale = "fixed",
  signal_y_ticks = "pretty",
  signal_track_height = 3,
  gene_track_height = 1
)

p_rnaseq_gene
```

Ribo-seq gene track:

```{r}
p_riboseq_gene <- plot_signal_gene(
  signal = riboseq,
  annotation = gp,
  gene_id = "GeneA",
  plot_type = "bar",
  strand = "auto",
  signal_palette = "Reds",
  signal_palette_direction = -1,
  signal_y_scale = "fixed",
  signal_y_ticks = "pretty",
  signal_track_height = 3,
  gene_track_height = 1
)

p_riboseq_gene
```

At a whole-gene genomic scale, thousands of bases are compressed into the available plot width, so a three-nucleotide pattern cannot be resolved visually even when the underlying RPF counts are frame-biased. The gene-level bar track is intended to show where translation signal occurs across CDS exons. Use the transcript-level `frame` view below, or a short genomic window, to inspect three-nucleotide periodicity.

The relative vertical space occupied by the signal and gene-model panels is controlled directly by `signal_track_height` and `gene_track_height`.

### Step 4. Plot RNA-seq and Ribo-seq tracks for a transcript

`plot_signal_transcript()` focuses on a single transcript. The RNA-seq example uses the standard bar representation in transcript coordinates:

```{r}
p_rnaseq_transcript <- plot_signal_transcript(
  signal = rnaseq,
  annotation = gp,
  transcript_id = "TxA1",
  coordinate = "transcript",
  plot_type = "bar",
  strand = "auto",
  signal_palette = "Blues",
  signal_palette_direction = -1,
  signal_y_scale = "fixed",
  signal_track_height = 3,
  gene_track_height = 1
)

p_rnaseq_transcript
```

For Ribo-seq, `plot_type = "frame"` maps P-site counts back onto the transcript reading frame. The demo is designed so that frame 0 contributes about 80% of total RPF counts, while frame 1 and frame 2 each contribute about 10%. Counts within each frame are deliberately heterogeneous rather than repeated at nearly identical heights. Discrete palettes such as `Set1` are assigned strictly in palette order: `frame0`, `frame1`, and `frame2` receive the first, second, and third palette colors, respectively:

```{r}
p_riboseq_transcript <- plot_signal_transcript(
  signal = riboseq,
  annotation = gp,
  transcript_id = "TxA1",
  coordinate = "transcript",
  plot_type = "frame",
  strand = "auto",
  frame_palette = "Set1",
  signal_track_height = 3,
  gene_track_height = 1
)

p_riboseq_transcript
```

Use `plot_type = "bar"` instead when the goal is to compare the genomic distribution of Ribo-seq P-site counts with the RNA-seq coverage representation:

```{r}
plot_signal_transcript(
  signal = riboseq,
  annotation = gp,
  transcript_id = "TxA1",
  coordinate = "transcript",
  plot_type = "bar",
  strand = "auto",
  signal_palette = "Reds",
  signal_palette_direction = -1
)
```

### Step 5. Plot RNA-seq and Ribo-seq together with `plot_tracks()`

Merge the two `BwgTrack` objects so RNA-seq and Ribo-seq can be displayed in one integrated track figure:

```{r}
signal_all <- merge_bwg(rnaseq, riboseq)

signal_all
summary_bwg(signal_all)
```

For the positive-strand `GeneA` locus, explicitly select the two plus-strand samples:

```{r}
p_signal_tracks <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  gene_id = "GeneA",
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  signal_palette = "Dark2",
  signal_palette_direction = 1,
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  heights = c(
    signal = 4,
    gene = 1,
    feature = 0.8,
    variant = 0.7
  )
)

p_signal_tracks
```

`plot_tracks()` uses its named `heights` vector to control panel proportions. The `signal` and `gene` entries play the same role as `signal_track_height` and `gene_track_height` in the dedicated signal plotting functions.

### Step 6. Plot RNA-seq and Ribo-seq tracks across a genomic region

`plot_signal_region()` is useful when the region contains several genes or when the target interval is not defined by a single gene/transcript.

For `bar` and `line` region plots, sample/group colors follow the sample/group level order and the standard RColorBrewer class order. Discrete colors are not selected by interpolating between the first and last palette colors. Heatmaps remain continuous gradients.

The following region contains the positive-strand `GeneA`, negative-strand `GeneB`, and positive-strand `GeneC`, making it useful for displaying both strand-specific RNA-seq tracks:

```{r}
p_rnaseq_region <- plot_signal_region(
  signal = rnaseq,
  annotation = gp,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  strand = "both",
  plot_type = "bar",
  signal_palette = "Blues",
  signal_palette_direction = -1,
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  signal_track_height = 4,
  gene_track_height = 1
)

p_rnaseq_region
```

Plot the Ribo-seq tracks over the same interval:

```{r}
p_riboseq_region <- plot_signal_region(
  signal = riboseq,
  annotation = gp,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  strand = "both",
  plot_type = "bar",
  signal_palette = "Reds",
  signal_palette_direction = -1,
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  signal_track_height = 4,
  gene_track_height = 1
)

p_riboseq_region
```

For large real-world signal files, use lazy `BwgTrack` access where appropriate so that only the requested genomic interval is retrieved during plotting.
