## VCF variant tracks

GeneTrackR stores VCF records in a `VariantTrack` object. This module follows one continuous workflow for the demo VCF:

1. read VCF data and inspect the `VariantTrack` object;
2. validate and summarize variants;
3. retrieve variants from a genomic region;
4. retrieve variants by gene or transcript;
5. filter variants by ID, type, or text pattern;
6. plot variant tracks.

The examples use the same `GeneA` / `TxA1` locus used elsewhere in the documentation so that variant, haplotype, LD, phenotype, and signal examples can be compared directly.

### Step 1. Read VCF data

Locate the demo VCF and annotation files:

```{r}
vcf_file <- system.file(
  "extdata",
  "gtr_demo_variants.vcf",
  package = "GeneTrackR"
)

gp_file <- system.file(
  "extdata",
  "gtr_demo.genePredExt",
  package = "GeneTrackR"
)

gp <- read_genepred(
  gp_file,
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)
```

Read the complete demo VCF into memory:

```{r}
vcf <- read_vcf(
  vcf_file,
  mode = "memory",
  keep_genotype = TRUE,
  verbose = FALSE,
  progress = FALSE
)

vcf

summary_vcf(vcf, chrom = 'chr1', start = 1000000, end = 8002500)
```

`read_vcf()` returns a `VariantTrack`. Standard VCF fields are normalized internally to columns such as `chrom`, `pos`, `variant_id`, `ref`, `alt`, `qual`, `filter`, `info`, and `variant_type`. Genotype columns are retained when `keep_genotype = TRUE`.

```{r}
head(vcf$data[, c(
  "chrom",
  "pos",
  "variant_id",
  "ref",
  "alt",
  "variant_type"
)])

vcf$meta$sample_names
```

Coordinates inside `VariantTrack` use **1-based genomic positions**, consistent with the VCF specification.

#### Large indexed VCF files

For a bgzip-compressed VCF with a `.tbi` or `.csi` index, `mode = "auto"` or `mode = "lazy"` avoids loading all records into memory. A lazy `VariantTrack` stores the source file and sample metadata first; variants are loaded only when a genomic region is requested.

```{r}
# large_vcf <- "/path/to/large.vcf.gz"
#
# vcf_lazy <- read_vcf(
#   large_vcf,
#   mode = "lazy"
# )
#
# vcf_lazy
#
# vcf_lazy_region <- retrieve_vcf(
#   vcf_lazy,
#   chrom = "chr1",
#   start = 12339001,
#   end = 12374500,
#   as = "VariantTrack"
# )
```

Lazy regional queries require an indexed VCF and `Rsamtools`. For small files such as the bundled demo VCF, memory mode is simpler.

### Step 2. Validate and summarize variants

Use `validate_vcf()` to check required fields, coordinates, allele fields, duplicated variant IDs, and consistency between `pos`, `start`, and `end`:

```{r}
vcf_validation <- validate_vcf(vcf)

vcf_validation$invalid_summary
vcf_validation$warnings
```

The validation result contains:

- `invalid_records`: individual records that failed a check;
- `invalid_summary`: number of invalid records by reason;
- `warnings`: non-record-specific validation messages.

Use `summary_vcf()` to summarize the current variant collection. By default, variants are counted by chromosome and inferred variant type:

```{r}
vcf_summary <- summary_vcf(vcf)
vcf_summary
```

The demo VCF intentionally contains SNPs and indels so the same file can be reused for variant visualization, haplotype analysis, and LD analysis.

### Step 3. Retrieve variants from a genomic region

`retrieve_vcf()` can return either a plain `data.table` or a new `VariantTrack`.

Retrieve the broader `GeneA` / `GeneB` / `GeneC` demonstration region as a table:

```{r}
region_table <- retrieve_vcf(
  vcf,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  as = "data.table",
  verbose = FALSE
)

region_table[, .(
  chrom,
  pos,
  variant_id,
  ref,
  alt,
  variant_type
)]
```

Use `as = "VariantTrack"` when the subset will be passed to another GeneTrackR function:

```{r}
region_variants <- retrieve_vcf(
  vcf,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  as = "VariantTrack",
  verbose = FALSE
)

region_variants
```

The distinction is important:

- `as = "data.table"` is convenient for inspection, filtering, and tabular analysis;
- `as = "VariantTrack"` preserves the GeneTrackR object interface for downstream plotting and analysis.

### Step 4. Retrieve variants by gene or transcript

When annotation is available, `retrieve_vcf()` can resolve genomic coordinates directly from a gene or transcript ID.

Retrieve variants inside the `GeneA` gene body:

```{r}
genea_variants <- retrieve_vcf(
  vcf,
  annotation = gp,
  gene_id = "GeneA",
  as = "VariantTrack",
  verbose = FALSE
)

genea_variants
```

The canonical demo design contains **11 variants within the GeneA gene body**.

Upstream and downstream flanking regions can be added explicitly:

```{r}
genea_extended <- retrieve_vcf(
  vcf,
  annotation = gp,
  gene_id = "GeneA",
  upstream = 1000,
  downstream = 500,
  strand_aware = TRUE,
  as = "VariantTrack",
  verbose = FALSE
)

genea_extended
```

For `GeneA`, the upstream extension additionally includes `varAup01`, which is intentionally designed with missing and heterozygous genotypes for edge-case testing.

Transcript-aware retrieval uses the same interface:

```{r}
txa1_variants <- retrieve_vcf(
  vcf,
  annotation = gp,
  transcript_id = "TxA1",
  as = "VariantTrack",
  verbose = FALSE
)

txa1_variants
```

For negative-strand genes or transcripts, `strand_aware = TRUE` makes `upstream` and `downstream` follow transcriptional direction rather than simply subtracting or adding genomic coordinates.

### Step 5. Filter variants by ID, type, or pattern

Retrieve one or more known variant IDs:

```{r}
selected_variants <- retrieve_vcf(
  vcf,
  variant_id = c("varA03", "varA04", "varA05"),
  as = "data.table",
  verbose = FALSE
)

selected_variants[, .(
  chrom,
  pos,
  variant_id,
  ref,
  alt,
  variant_type
)]
```

Filter by inferred variant type:

```{r}
genea_indels <- retrieve_vcf(
  genea_variants,
  variant_type = c("INS", "DEL"),
  as = "data.table",
  verbose = FALSE
)

genea_indels[, .(
  pos,
  variant_id,
  ref,
  alt,
  variant_type
)]
```

`pattern` searches `variant_id`, `REF`, `ALT`, `INFO`, and `variant_type`. For example, the demo high-LD variants contain `high_ld` in the INFO role field:

```{r}
high_ld_variants <- retrieve_vcf(
  vcf,
  pattern = "high_ld",
  fixed = TRUE,
  as = "VariantTrack",
  verbose = FALSE
)

high_ld_variants
```

These filters can be combined with genomic, gene, or transcript queries when a more specific subset is needed.

### Step 6. Plot variant tracks

`plot_variant()` returns a ggplot object directly. The GeneA region contains SNP, insertion, and deletion examples, so it is useful for demonstrating variant-type colors.

```{r}
p_genea_variants <- plot_variant(
  genea_variants,
  color_by = "variant_type",
  label_by = "variant_id",
  variant_shape = "lollipop",
  variant_palette = "Set1",
  point_size = 2.5,
  text_size = 12
)

p_genea_variants
```

For a broader genomic region, use the previously retrieved `VariantTrack`:

```{r}
p_region_variants <- plot_variant(
  region_variants,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  color_by = "variant_type",
  label_by = "none",
  variant_shape = "rug",
  variant_palette = "Set1",
  text_size = 12
)

p_region_variants
```

The main plotting controls are:

- `color_by`: `"variant_type"`, `"filter"`, or `"none"`;
- `label_by`: variant ID, type, REF, ALT, or no labels;
- `variant_shape`: `"lollipop"`, `"point"`, or `"rug"`;
- `variant_palette` / `variant_colors`: automatic palette or explicit colors.

Variant subsets created here are reused directly by the later haplotype, phenotype, LD, and integrated-track modules.

VCF export is covered separately in the **Export variants and analysis results** module. The current `write_vcf()` interface writes standardized site-level VCF fields; the original genotype-containing VCF should be retained as the archival source when genotype round-trip preservation is required.
