## Variant track module

### Read VCF into memory

```{r}
vcf <- read_vcf(vcf_file, mode = "memory")
vcf
summary_vcf(vcf)
```

### Lazy indexed VCF access

For large `vcf.gz` files with `.tbi` or `.csi` indexes, use lazy mode. This reads the header and sample names first, then reads variants only for requested regions.

```{r}
large_vcf <- "/path/to/large.vcf.gz"

# For a real bgzip-compressed VCF with a .tbi/.csi index:
# vcf_lazy <- read_vcf(large_vcf, mode = "lazy")
# vcf_region <- retrieve_vcf(
#   vcf_lazy,
#   chrom = "chr1",
#   start = 12339001,
#   end = 12374500,
#   as = "VariantTrack"
# )
```

If no index is available, use memory mode or create a tabix index before lazy retrieval.

### Retrieve and plot variants

```{r}
vcf_region <- retrieve_vcf(
  vcf,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  as = "VariantTrack"
)

vcf_region
head(vcf_region$data)

plot_variant(
  vcf,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  color_by = "variant_type"
)
```

`retrieve_vcf()` returns a `data.table` by default. Set `as = "VariantTrack"` when the retrieved subset itself should be passed to functions that require a `VariantTrack`.
