## Browser-like combined tracks

`plot_tracks()` combines gene models, signal tracks, BED feature tracks, and VCF variant tracks.
It returns the assembled ggplot/patchwork figure directly rather than wrapping it in a result list.

```{r}
plot_tracks(
  annotation = gp,
  signal = rnaseq,
  gene_id = "GeneA",
  signal_type = "bar"
)
```

Customize gene model colors in combined tracks:

```{r}
plot_tracks(
  annotation = gp,
  signal = rnaseq,
  gene_id = "GeneA",
  signal_type = "bar",
  gene_palette = "Set2",
  gene_border_color = "black"
)

plot_tracks(
  annotation = gp,
  signal = rnaseq,
  gene_id = "GeneA",
  gene_colors = c(
    CDS = "#1b9e77",
    UTR = "#a6d854",
    exon = "#7570b3"
  )
)
```

Control the complete track style with standardized parameters:

```{r}
plot_tracks(
  annotation = gp,
  signal = rnaseq,
  gene_id = "GeneA",
  signal_type = "bar",
  signal_transform = "sqrt",
  signal_y_scale = "fixed",
  signal_y_limits = c(0, 20),
  signal_alpha = 0.80,
  signal_bar_width = 0.85,
  plot_theme = "classic",
  show_panel_border = FALSE
)
```

Add feature and variant tracks:

```{r}
plot_tracks(
  annotation = gp,
  signal = signal_all,
  features = features,
  variants = vcf,
  chrom = "chr1",
  start = 12339001,
  end = 12374500,
  signal_type = "bar"
)
```
