## Browser-like combined tracks

`plot_tracks()` assembles gene models, genomic signal, interval features, and variants into one genome-browser-like figure with a shared genomic x-axis. It is intended for integrated inspection of a locus rather than for replacing the dedicated plotting functions for each data type.

This module follows one continuous workflow:

1. prepare annotation, signal, feature, and variant tracks;
2. build a gene-centered browser view;
3. build a transcript-centered browser view;
4. inspect an explicit genomic region;
5. add BED features and VCF variants;
6. control layout, panel heights, highlighting, and styling;
7. choose an appropriate browser workflow for large real-world datasets.

The examples use the same `GeneA` / `TxA1` region used throughout the other GeneTrackR modules.

### Step 1. Prepare browser-track inputs

Locate the bundled demo files:

```{r}
gp_file <- system.file(
  "extdata",
  "gtr_demo.genePredExt",
  package = "GeneTrackR"
)

signal_files <- system.file(
  "extdata",
  c(
    "gtr_demo_rnaseq_plus.bedgraph",
    "gtr_demo_rnaseq_minus.bedgraph",
    "gtr_demo_riboseq_plus.bedgraph",
    "gtr_demo_riboseq_minus.bedgraph"
  ),
  package = "GeneTrackR"
)

feature_file <- system.file(
  "extdata",
  "gtr_demo_features.bed",
  package = "GeneTrackR"
)

vcf_file <- system.file(
  "extdata",
  "gtr_demo_variants.vcf",
  package = "GeneTrackR"
)
```

Read the annotation, four strand-specific signal tracks, BED features, and VCF variants:

```{r}
gp <- read_genepred(
  gp_file,
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)

signal_all <- read_bwg(
  signal_files,
  format = "bedgraph",
  sample_names = c(
    "RNA_seq_plus",
    "RNA_seq_minus",
    "Ribo_seq_plus",
    "Ribo_seq_minus"
  ),
  strand = c("+", "-", "+", "-"),
  mode = "memory",
  verbose = FALSE
)

features <- read_bed(
  feature_file,
  verbose = FALSE,
  progress = FALSE
)

variants <- read_vcf(
  vcf_file,
  mode = "memory",
  keep_genotype = TRUE,
  verbose = FALSE,
  progress = FALSE
)
```

The four inputs play different roles in `plot_tracks()`:

- `annotation`: gene/transcript/exon structure used for the gene-model panel;
- `signal`: optional `BwgTrack` containing one or more quantitative tracks;
- `features`: optional `FeatureTrack` containing BED/GFF/GTF-style intervals;
- `variants`: optional `VariantTrack` containing VCF-derived variants.

General GeneTrackR plotting palette arguments default to `"Paired"`. The deliberate exception is `plot_ld_block()`, whose `color_palette` defaults to the sequential `"Reds"` palette for LD heatmaps. Supply another RColorBrewer palette explicitly when a different color scheme is required.

`plot_tracks()` accepts **exactly one genomic locator** per call:

- `gene_id`;
- `transcript_id`;
- or the complete `chrom` + `start` + `end` combination.

Do not supply more than one locator type in the same call.

### Step 2. Build a gene-centered browser view

Start with the annotation alone. When `gene_id` is supplied, GeneTrackR derives the chromosome and plotting interval from the complete gene locus:

```{r}
p_browser_gene_model <- plot_tracks(
  annotation = gp,
  gene_id = "GeneA",
  gene_palette = "Paired",
  direction_mode = "transcript"
)

p_browser_gene_model
```

Add RNA-seq and Ribo-seq signal over the same locus. `GeneA` is on the positive strand, so select the two positive-strand samples explicitly:

```{r}
p_browser_gene <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  gene_id = "GeneA",
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  signal_palette = "Paired",
  signal_palette_direction = 1,
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  heights = c(
    signal = 4,
    gene = 1,
    feature = 0.8,
    variant = 0.7
  )
)

p_browser_gene
```

For discrete signal tracks, colors follow the selected sample/group order and the standard palette order. Unlike `plot_signal_gene()`, `plot_tracks()` does not use `strand = "auto"`. Use `+` or `-` when a strand-specific locus is known, `both` when both strand classes should be shown, or `ignore` when signal strand metadata should not be used for filtering.

### Step 3. Build a transcript-centered browser view

Use `transcript_id` when the plotting interval should correspond to one transcript rather than the complete gene locus:

```{r}
p_browser_transcript <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  transcript_id = "TxA1",
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  ribo_signal_type = "auto",
  frame_palette = "Paired",
  signal_y_scale = "free",
  collapse = "none",
  direction_mode = "transcript",
  heights = c(
    signal = 4,
    gene = 1,
    feature = 0.8,
    variant = 0.7
  )
)

p_browser_transcript
```

When `transcript_id` is supplied, `ribo_signal_type = "auto"` detects Ribo-seq/RPF-like sample IDs and renders those panels in `frame` mode by default, while RNA-seq panels remain standard browser signal tracks. This makes the difference between RNA-seq coverage and Ribo-seq frame periodicity more obvious without requiring a separate manual step.

The signal panel remains in **genomic coordinates** so that it stays aligned with the other browser tracks. For region-level browser plots, Ribo-seq defaults back to standard genomic signal tracks because frame rendering requires one specific transcript model.

### Step 4. Inspect an explicit genomic region

For a multi-gene interval or a locus not defined by one annotation ID, provide explicit genomic coordinates. The demo interval below contains `GeneA`, `GeneB`, and `GeneC` and therefore includes both positive- and negative-strand signal:

```{r}
browser_chrom <- "chr1"
browser_start <- 12339001
browser_end <- 12374500

p_browser_region <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  chrom = browser_chrom,
  start = browser_start,
  end = browser_end,
  samples = c(
    "RNA_seq_plus",
    "RNA_seq_minus",
    "Ribo_seq_plus",
    "Ribo_seq_minus"
  ),
  strand = "both",
  signal_type = "bar",
  signal_palette = "Paired",
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  collapse = "none",
  direction_mode = "gene",
  heights = c(
    signal = 5,
    gene = 1.2,
    feature = 0.8,
    variant = 0.7
  )
)

p_browser_region
```

Use `signal_y_scale = "free"` when assays have substantially different count ranges, as in the RNA-seq and Ribo-seq demo. Use `fixed` when the signal panels should share the same y-axis range for direct quantitative comparison.

### Step 5. Add BED features and VCF variants

Pass `FeatureTrack` and `VariantTrack` objects directly to `features` and `variants`. `plot_tracks()` retrieves only records overlapping the selected browser interval:

```{r}
p_browser_complete <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  features = features,
  variants = variants,
  chrom = browser_chrom,
  start = browser_start,
  end = browser_end,
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  signal_palette = "Paired",
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  gene_palette = "Paired",
  feature_color_by = "auto",
  feature_max_legend_levels = 5,
  variant_palette = "Paired",
  variant_colors = c(
    SNP = "#1F78B4",
    INS = "#33A02C",
    DEL = "#E31A1C"
  ),
  direction_mode = "gene",
  heights = c(
    signal = 4,
    gene = 1.2,
    feature = 0.9,
    variant = 0.8
  )
)

p_browser_complete
```

For BED-like interval tracks, `feature_color_by = "auto"` tries to choose a compact and informative attribute for the legend. In the bundled demo BED file, names such as `GeneA_promoter|promoter` and `GeneB_enhancer|enhancer` are simplified to a small set of feature groups (for example promoter, enhancer, candidate region, and QTL-like intervals), and the legend is automatically capped to at most five groups. This is usually more informative than a single `BED` legend entry, while still keeping the browser legend manageable.

Variant colors are controlled independently with `variant_palette` and `variant_colors`. `variant_palette = "Paired"` is the default. For stable biological categories, a named vector such as `c(SNP = ..., INS = ..., DEL = ...)` is preferable because the same variant type then retains the same color across browser regions.

The combined figure uses the following default panel order when `layout = "signal_top"`:

1. signal;
2. feature track(s);
3. variant track(s);
4. gene model.

With `layout = "gene_top"`, the gene model is moved to the top, followed by signal, feature, and variant panels.

`features` and `variants` can also be named lists when several independent interval or variant collections should be drawn as separate panels:

```{r}
# feature_tracks <- list(
#   Regulatory = regulatory_features,
#   QTL = qtl_features
# )
#
# variant_tracks <- list(
#   Natural = natural_variants,
#   Candidate = candidate_variants
# )
#
# plot_tracks(
#   annotation = gp,
#   features = feature_tracks,
#   variants = variant_tracks,
#   chrom = browser_chrom,
#   start = browser_start,
#   end = browser_end
# )
```

Each list entry becomes a separate panel. The `feature` and `variant` entries of `heights` are reused for every panel of the corresponding class.

### Step 6. Control layout, heights, highlighting, and style

Panel proportions are controlled through the named `heights` vector. Increase `signal` when quantitative tracks need more vertical space, or increase `gene`, `feature`, and `variant` when structural tracks are crowded.

The following example moves the gene model to the top and highlights the small LD demonstration interval shared with the LD workflow:

```{r}
ld_highlight <- data.frame(
  start = 12342620,
  end = 12343180
)

p_browser_styled <- plot_tracks(
  annotation = gp,
  signal = signal_all,
  features = features,
  variants = variants,
  chrom = browser_chrom,
  start = browser_start,
  end = browser_end,
  samples = c("RNA_seq_plus", "Ribo_seq_plus"),
  strand = "+",
  signal_type = "bar",
  signal_palette = "Paired",
  signal_transform = "sqrt",
  signal_y_scale = "free",
  signal_y_ticks = "pretty",
  signal_alpha = 0.85,
  signal_bar_width = 0.9,
  gene_palette = "Paired",
  gene_border_color = "black",
  feature_color_by = "auto",
  feature_max_legend_levels = 5,
  highlight = ld_highlight,
  layout = "gene_top",
  heights = c(
    signal = 4,
    gene = 1.5,
    feature = 0.9,
    variant = 0.8
  ),
  plot_theme = "classic",
  show_panel_border = FALSE,
  text_size = 14
)

p_browser_styled
```

`highlight` uses genomic coordinates and is currently applied to the signal and gene-model panels. It is useful for marking a candidate interval while preserving the common browser x-axis.

Gene-model styling is controlled by `gene_palette`, `gene_colors`, `gene_border_color`, `cds_height`, `utr_height`, `direction_mode`, `label_position`, and `label_by`. Signal styling is controlled by the corresponding `signal_*` arguments. Feature styling is controlled by `feature_color_by`, `feature_palette`, `feature_colors`, `feature_border_color`, and `feature_max_legend_levels`. Variant styling in the integrated browser is controlled by `variant_palette` and `variant_colors`.

The integrated browser intentionally uses simplified feature and variant rendering with labels hidden. When detailed interval labels, variant labels, or more detailed track-specific customization are required, first inspect them with `plot_feature_track()`, `plot_variant()`, or `plot_signal_transcript()`, then use `plot_tracks()` for the compact integrated overview.

### Step 7. Use browser tracks efficiently on real datasets

For large datasets, avoid loading genome-wide signal or VCF records when regional access is available:

- use lazy `BwgTrack` access for indexed bigWig or supported large signal inputs;
- use `read_vcf(mode = "lazy")` with a bgzip-compressed, indexed VCF for regional variant retrieval;
- keep the browser interval biologically focused rather than plotting an entire chromosome with base-level Ribo-seq or dense variants;
- preselect relevant signal samples with `samples` before plotting;
- use dedicated functions such as `plot_signal_transcript()`, `plot_variant()`, or `plot_feature_track()` when one track requires detailed analysis beyond the compact browser representation.

`plot_tracks()` returns the assembled ggplot/patchwork figure directly. Figure export is covered in the export workflow rather than duplicated here.
