## Haplotype extraction

GeneTrackR now separates gene/transcript-based and region-based haplotype extraction.

### Gene or transcript haplotypes

```{r}
hap_gene <- hap_gene_variant(
  vcf,
  annotation = gp,
  gene_id = "GeneA",
  genotype_mode = "string",
  min_variant_number = 1
)

hap_tx <- hap_gene_variant(
  vcf,
  annotation = gp,
  transcript_id = "TxA1",
  genotype_mode = "code",
  min_variant_number = 1
)
```

Include upstream/downstream variants around a gene or transcript:

```{r}
hap_gene_ext <- hap_gene_variant(
  vcf,
  annotation = gp,
  gene_id = "GeneA",
  upstream = 1000,
  downstream = 1000,
  strand_aware = TRUE,
  genotype_mode = "string",
  min_variant_number = 1
)
```

### Region haplotypes

```{r}
hap_region <- hap_region_variant(
  vcf,
  chrom = "chr1",
  start = 12339700,
  end = 12352000,
  genotype_mode = "code",
  min_variant_number = 1
)
```

`hap_variant()` is retained as a compatibility wrapper, but new code should prefer `hap_gene_variant()` and `hap_region_variant()`.

### Inspect haplotype tables

```{r}
hap_gene$region
hap_gene$variants
hap_gene$haplotypes
hap_gene$sample_haplotypes
```

`hap_gene` is a `HapVariant`. Downstream haplotype plots and phenotype/refinement functions consume this object directly; there is no need to reconstruct genotype tables manually.

## Haplotype-variant plot

`plot_hap_variant()` draws a gene model, variant markers, connector lines, and a genotype table.

```{r}
hap_variant_figure <- plot_hap_variant(
  hap_gene,
  annotation = gp,
  min_hap_samples = 3,
  show_reference_row = TRUE,
  table_x_angle = 90
)

hap_variant_figure
```

Customize table and variant colors:

```{r}
plot_hap_variant(
  hap_gene,
  annotation = gp,
  min_hap_samples = 3,
  table_palette = "RdBu",
  table_alpha = 0.6,
  variant_palette = "Set1",
  genotype_text_size = 3
)
```
