## Phenotype association workflow

GeneTrackR keeps phenotype data as a simple sample-level table and joins it to haplotype or variant genotypes by `sample_id`. The recommended workflow is to inspect phenotype quality first, verify sample matching explicitly, then perform haplotype- or variant-based association tests and keep the returned statistical tables together with the figure.

This module follows one continuous workflow:

1. prepare phenotype, VCF, annotation, and the standard GeneA haplotypes;
2. summarize phenotype types, missingness, and distributions;
3. verify sample matching before association analysis;
4. test haplotype-phenotype associations;
5. inspect multiple traits and statistical-test settings;
6. control phenotype-association plot appearance;
7. test a single variant against phenotype;
8. pass the same phenotype and haplotype objects to refinement and variant-effect workflows.

The bundled phenotype data contain four numeric traits and one categorical trait. Their designed roles are:

| Trait | Type | Demo role |
| --- | --- | --- |
| `seed_weight` | numeric | strong four-level GeneA haplotype effect with nearest genotype pairs remaining phenotypically closer |
| `plant_height` | numeric | moderate four-level GeneA haplotype effect with the same hierarchical ordering |
| `protein_content` | numeric | strong `varA03` effect aligned with the `Hap1/Hap2` versus `Hap3/Hap4` genotype split |
| `flowering_time` | numeric | negative control for GeneA haplotypes |
| `flower_color` | categorical | categorical phenotype example |

### Step 1. Prepare phenotype and genotype inputs

Locate the bundled phenotype, VCF, and annotation files:

```{r}
pheno_file <- system.file(
  "extdata",
  "gtr_demo_pheno.tsv",
  package = "GeneTrackR"
)

vcf_file <- system.file(
  "extdata",
  "gtr_demo_variants.vcf",
  package = "GeneTrackR"
)

gp_file <- system.file(
  "extdata",
  "gtr_demo.genePredExt",
  package = "GeneTrackR"
)
```

Read the three inputs and reconstruct the same GeneA gene-body haplotypes used in the haplotype workflow:

```{r}
pheno <- read_pheno(
  pheno_file,
  verbose = FALSE,
  progress = FALSE
)

vcf <- read_vcf(
  vcf_file,
  mode = "memory",
  keep_genotype = TRUE,
  verbose = FALSE,
  progress = FALSE
)

gp <- read_genepred(
  gp_file,
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)

hap_gene <- hap_gene_variant(
  vcf = vcf,
  annotation = gp,
  gene_id = "GeneA",
  genotype_mode = "string"
)
```

`read_pheno()` returns a `data.table`. By default, the first column is treated as the sample column and renamed to `sample_id`. If another column stores sample IDs, specify it explicitly:

```{r}
# pheno <- read_pheno(
#   "phenotype.tsv",
#   sample_col = "Taxa"
# )
```

Sample IDs must be unique. Duplicate IDs are rejected because they would make genotype-phenotype matching ambiguous.

### Step 2. Summarize phenotype quality and distributions

Start with `summary_pheno()` before any association test:

```{r}
pheno_summary <- summary_pheno(pheno)
pheno_summary
```

The summary reports:

- numeric versus categorical trait type;
- total sample number;
- missing number and missing rate;
- number of unique non-missing values;
- minimum, mean, median, and maximum for numeric traits.

Plot numeric and categorical phenotype distributions directly from the phenotype table:

```{r}
pheno_figure <- plot_pheno(
  pheno,
  traits = c(
    "seed_weight",
    "protein_content",
    "plant_height",
    "flowering_time",
    "flower_color"
  )
)

pheno_figure
```

`plot_pheno()` returns the figure directly. Numeric traits are shown as histograms, while categorical traits are shown as count bars. When both trait classes are requested, the result is a combined patchwork figure.

`plot_hap_pheno()` and `plot_variant_pheno()` are currently intended for **numeric phenotype traits**. Categorical traits such as `flower_color` should therefore be summarized or plotted with `summary_pheno()` / `plot_pheno()` unless a categorical association method is implemented separately.

### Step 3. Verify sample matching explicitly

Phenotype rows do not need to follow the VCF or haplotype sample order. GeneTrackR joins genotype and phenotype records by sample ID.

The demo phenotype table intentionally uses a different row order from the genotype data. Check the overlap before testing:

```{r}
hap_sample_ids <- as.character(hap_gene$sample_haplotypes$sample_id)
pheno_sample_ids <- as.character(pheno$sample_id)

sample_match <- data.frame(
  phenotype_samples = length(unique(pheno_sample_ids)),
  haplotype_samples = length(unique(hap_sample_ids)),
  matched_samples = length(intersect(pheno_sample_ids, hap_sample_ids)),
  phenotype_only = length(setdiff(pheno_sample_ids, hap_sample_ids)),
  haplotype_only = length(setdiff(hap_sample_ids, pheno_sample_ids))
)

sample_match
```

For the bundled demo, all 36 phenotype samples match the 36 GeneA haplotype samples even though their row orders differ.

In real datasets, samples present on only one side are not available for the association test. Checking `setdiff()` before plotting is therefore recommended, especially after genotype missingness filters or phenotype QC have removed samples.

### Step 4. Test the primary haplotype-phenotype association

`seed_weight` is the main deterministic GeneA haplotype phenotype in the demo. Test the four GeneA haplotypes with `plot_hap_pheno()`:

```{r}
hap_seed <- plot_hap_pheno(
  hap = hap_gene,
  phenotype = pheno,
  traits = "seed_weight",
  min_hap_samples = 3,
  plot_type = "violin_boxplot",
  test_method = "t.test",
  p_adjust = "BH",
  p_value_type = "adjusted",
  p_label = "stars",
  show_signif_only = TRUE,
  fill_palette = "Paired"
)

hap_seed$figure
```

`plot_hap_pheno()` returns a `GeneTrackRPhenoPlot`, not only a ggplot. Keep the complete result because it contains the statistical output used to draw the figure:

```{r}
hap_seed$pvalue
hap_seed$summary
hap_seed$bracket
```

The main components are:

| Component | Content |
| --- | --- |
| `$figure` | phenotype distribution figure |
| `$pvalue` | all pairwise tests, including raw and adjusted p-values |
| `$summary` | sample number, median, mean, minimum, and maximum by trait and haplotype |
| `$bracket` | comparisons actually prepared for display on the figure |
| `$plot_data` | matched long-format phenotype data used for plotting |

Printing the complete `GeneTrackRPhenoPlot` also prints its `$figure`, but use the named components when saving figures or exporting statistical tables.

Haplotype groups are ordered by retained sample number. The x-axis label includes the group sample count, for example `Hap1 (9)`.

### Step 5. Compare traits and choose the statistical settings

The demo provides three useful haplotype-level behaviors in one dataset:

- `seed_weight`: strong four-level differences, while the genotype-nearest pairs remain phenotypically closer than cross-cluster pairs;
- `plant_height`: the same hierarchical pattern with a more moderate effect scale;
- `flowering_time`: negative control without a designed GeneA haplotype effect.

Plot them together while preserving the requested trait order:

```{r}
hap_multi <- plot_hap_pheno(
  hap = hap_gene,
  phenotype = pheno,
  traits = c(
    "seed_weight",
    "plant_height",
    "flowering_time"
  ),
  min_hap_samples = 3,
  test_method = "t.test",
  p_adjust = "BH",
  p_value_type = "adjusted",
  p_label = "stars",
  facet_ncol = 3,
  fill_palette = "Paired"
)

hap_multi$figure
hap_multi$pvalue
```

Three pairwise test methods are available:

- `t.test`: compares group means and is appropriate for approximately continuous, reasonably behaved traits;
- `wilcox.test`: rank-based alternative that is less dependent on normality assumptions;
- `ks.test`: tests whether the overall distributions differ, not specifically their means or medians.

For example:

```{r}
hap_seed_wilcox <- plot_hap_pheno(
  hap = hap_gene,
  phenotype = pheno,
  traits = "seed_weight",
  min_hap_samples = 3,
  test_method = "wilcox.test",
  p_adjust = "BH",
  p_value_type = "adjusted"
)

hap_seed_wilcox$pvalue
```

`$pvalue` always retains both `p_value` and `p_adj`. The argument `p_value_type` controls which of these is used to filter and label significance brackets; it does not remove the other statistic from the returned table.

With four haplotypes there are six pairwise comparisons per trait, so adjusted p-values are generally preferable for the main interpretation. `p_adjust = "BH"` is the default multiple-testing adjustment used in these examples.

`show_signif_only = TRUE` only controls bracket display. Non-significant comparisons remain available in `$pvalue`.

### Step 6. Control phenotype-association plot appearance

General GeneTrackR plotting palettes default to `Paired`; `plot_ld_block()` is the deliberate package-level exception and defaults to `Reds`. `plot_hap_pheno()` maps fill colors according to the **median phenotype value within each trait**, so the colors are used as an ordered visual encoding rather than as permanent haplotype identities.

A more detailed plot can be requested as follows:

```{r}
hap_seed_styled <- plot_hap_pheno(
  hap = hap_gene,
  phenotype = pheno,
  traits = "seed_weight",
  min_hap_samples = 3,
  plot_type = "violin_boxplot",
  test_method = "t.test",
  p_adjust = "BH",
  p_value_type = "adjusted",
  p_label = "both",
  show_points = TRUE,
  show_outliers = FALSE,
  fill_palette = "Paired",
  fill_alpha = 0.75,
  x_text_angle = 90,
  text_size = 14
)

hap_seed_styled$figure
```

Use `plot_type = "violin"`, `"boxplot"`, or `"violin_boxplot"` depending on whether the distribution shape, compact summary, or both are most useful.

If fixed colors for specific haplotype IDs are required, supply a named vector instead of relying on the median-based palette assignment:

```{r}
hap_seed_fixed_colors <- plot_hap_pheno(
  hap = hap_gene,
  phenotype = pheno,
  traits = "seed_weight",
  min_hap_samples = 3,
  fill_colors = c(
    Hap1 = "#A6CEE3",
    Hap2 = "#1F78B4",
    Hap3 = "#B2DF8A",
    Hap4 = "#33A02C"
  )
)

hap_seed_fixed_colors$figure
```

Use fixed haplotype colors only after confirming the actual `hap_id` values in `hap_gene$haplotypes`; haplotype numbering is specific to the variant set used to construct each `HapVariant`.

### Step 7. Test a single variant against phenotype

`plot_variant_pheno()` is the single-variant counterpart of `plot_hap_pheno()`. The demo variant `varA03` follows the same genotype-cluster split used by refinement: `Hap1/Hap2` carry the REF class, `Hap3/Hap4` carry the ALT class, and ALT carriers have higher `protein_content` values.

Select the variant by ID and use the compact binary REF/ALT representation:

```{r}
variant_protein <- plot_variant_pheno(
  variant = vcf,
  phenotype = pheno,
  variant_id = "varA03",
  traits = "protein_content",
  genotype_mode = "code",
  min_group_samples = 3,
  test_method = "t.test",
  p_adjust = "BH",
  p_value_type = "adjusted",
  p_label = "stars",
  fill_palette = "Paired"
)

variant_protein$figure
variant_protein$pvalue
variant_protein$summary
variant_protein$variant_data
```

For `genotype_mode = "code"`:

- `0` means reference-only genotype;
- `1` means at least one alternate allele is present.

This is a binary REF/ALT-presence representation, not allele dosage. Use `genotype_mode = "string"` when allele labels are more informative:

```{r}
variant_protein_string <- plot_variant_pheno(
  variant = vcf,
  phenotype = pheno,
  variant_id = "varA03",
  traits = "protein_content",
  genotype_mode = "string",
  min_group_samples = 3,
  p_value_type = "adjusted"
)

variant_protein_string$figure
```

A variant can also be selected by exact genomic position:

```{r}
variant_protein_position <- plot_variant_pheno(
  variant = vcf,
  phenotype = pheno,
  chrom = "chr1",
  pos = 12342550,
  traits = "protein_content",
  genotype_mode = "code",
  min_group_samples = 3,
  p_value_type = "adjusted"
)

variant_protein_position$variant_data
```

The selected locator must resolve to exactly one variant. If a region contains several variants, provide `variant_id` or an exact chromosome-position locator.

`plot_variant_pheno()` returns the same `GeneTrackRPhenoPlot` structure as `plot_hap_pheno()`, with one additional component: `$variant_data`, containing the selected variant record.

### Step 8. Pass the same objects downstream

Keep the same `hap_gene` and `pheno` objects for phenotype-guided refinement and variant-effect prioritization. Rebuilding the haplotypes with a different region or missing-data rule would change the biological groups being tested.

```{r}
# Haplotype refinement (11-haplotype-refinement.qmd)
# refined <- refine_haplotype(
#   hap_gene,
#   phenotype = pheno,
#   traits = "protein_content",
#   min_hap_samples = 3,
#   effect_threshold = 0.5
# )

# Variant-effect prioritization (12-variant-effect.qmd)
# effect <- plot_variant_effect(
#   hap_gene,
#   phenotype = pheno,
#   traits = "protein_content",
#   min_group_samples = 3
# )
```

The resulting workflow keeps one consistent chain:

`VCF + annotation -> HapVariant -> phenotype association -> refinement / variant-effect analysis`.

For reproducible analysis, export `$pvalue`, `$summary`, and the matched sample-level `$plot_data` together with `$figure` rather than saving only the visual result.
