## Phenotype input and summary

The phenotype table should contain sample/taxa IDs in the first column or a named sample column. Each additional column is treated as one trait.

```{r}
pheno <- read_pheno(pheno_file)

summary_pheno(pheno)

plot_pheno(
  pheno,
  traits = c("plant_height", "seed_weight")
)
```

`read_pheno()` returns a `data.table`; `plot_pheno()` returns a figure directly.

## Haplotype-phenotype association

`plot_hap_pheno()` compares phenotype distributions among haplotype groups. It returns both the figure and the p-value table.

```{r}
hap_res <- plot_hap_pheno(
  hap = hap_gene,
  phenotype = pheno,
  traits = "plant_height",
  min_hap_samples = 3,
  test_method = "t.test",
  p_value_type = "raw",
  p_label = "stars"
)

hap_res$figure
hap_res$pvalue
hap_res$summary
```

Use Wilcoxon or KS tests:

```{r}
plot_hap_pheno(
  hap = hap_gene,
  phenotype = pheno,
  traits = "seed_weight",
  test_method = "wilcox.test",
  min_hap_samples = 3
)$figure
```

Multiple traits preserve the requested order and are arranged over multiple facet columns:

```{r}
plot_hap_pheno(
  hap = hap_gene,
  phenotype = pheno,
  traits = c(
    "seed_weight",
    "plant_height",
    "protein_content",
    "flowering_time"
  ),
  facet_ncol = 2,
  min_hap_samples = 3
)$figure
```

## Single-variant phenotype association

`plot_variant_pheno()` is the single-variant version of `plot_hap_pheno()`. It groups samples by genotype or allele state at one variant.

```{r}
variant_res <- plot_variant_pheno(
  variant = vcf,
  phenotype = pheno,
  variant_id = "varA03",
  traits = "protein_content",
  genotype_mode = "string",
  min_group_samples = 3
)

variant_res$figure
variant_res$pvalue
variant_res$variant_data
```

You can also select a variant by genomic position:

```{r}
plot_variant_pheno(
  variant = vcf,
  phenotype = pheno,
  chrom = "chr1",
  pos = 12342550,
  traits = "protein_content",
  genotype_mode = "code",
  min_group_samples = 3
)$figure
```
