## Linkage disequilibrium analysis and visualization

GeneTrackR separates linkage disequilibrium analysis into two stages. `compute_ld_block()` calculates pairwise LD from VCF genotypes and stores the complete result in an `LDTrack`; `plot_ld_block()` then adds a triangular heatmap and optionally a gene/variant region track to the same object. Keeping calculation and plotting separate makes the numerical LD table, matrix, region definition, sample settings, and final figure available for downstream analysis.

The bundled demo contains three deterministic LD scenarios:

- `varLD01`-`varLD06`: six variants with the same genotype pattern, giving 15 pairwise comparisons with `r2 = 1`;
- `varLow01`-`varLow04`: four variants designed to have little linkage, with most pairwise `r2 = 0` and a maximum of approximately `0.111`;
- `varPair01`-`varPair02`: exactly two variants, retained to test the single-diamond LD plot.

This module follows nine steps:

1. prepare genotype-rich VCF data and gene annotation;
2. compute the primary six-variant high-LD example;
3. inspect the `LDTrack` object and pairwise statistics;
4. inspect the symmetric LD matrix and define descriptive pair filters;
5. draw the triangular LD heatmap;
6. add gene structure, natural-variant markers, and connector lines;
7. compare high-LD, low-LD, and two-variant regions;
8. control samples, variant types, LD methods, and retained genotype matrices;
9. interpret LD results and pass a selected region to haplotype analysis.

### Step 1. Prepare VCF genotypes and annotation

Locate and read the same demo VCF and GenePred annotation used in the variant and haplotype workflows:

```{r}
vcf_file <- system.file(
  "extdata",
  "gtr_demo_variants.vcf",
  package = "GeneTrackR"
)

gp_file <- system.file(
  "extdata",
  "gtr_demo.genePredExt",
  package = "GeneTrackR"
)

vcf <- read_vcf(
  vcf_file,
  mode = "memory",
  keep_genotype = TRUE,
  verbose = FALSE,
  progress = FALSE
)

gp <- read_genepred(
  gp_file,
  format = "genePredExt",
  verbose = FALSE,
  progress = FALSE
)
```

LD calculation requires VCF genotype columns. If the VCF was previously read with `keep_genotype = FALSE`, read it again with `keep_genotype = TRUE` before calling `compute_ld_block()`.

For large bgzip-compressed and indexed VCF files, a lazy `VariantTrack` can be used, but lazy LD queries require a complete `chrom`, `start`, and `end` interval so only the requested region is retrieved.

### Step 2. Compute the primary high-LD example

The main demo region contains exactly six designed high-LD variants:

```{r}
ld_high <- compute_ld_block(
  vcf,
  chrom = "chr1",
  start = 12342620,
  end = 12343180,
  method = "r2",
  min_pair_samples = 3,
  verbose = FALSE
)

ld_high
```

The expected result is:

- 6 retained variants;
- 15 unique variant pairs (`6 * 5 / 2`);
- 36 available samples per pair;
- `r2 = 1` for every pair.

Check the retained variants before interpreting the heatmap:

```{r}
ld_high$variants
```

`compute_ld_block()` sorts variants by genomic position and assigns a `variant_index`. The plotting functions use this same order, so the pairwise table, matrix, region markers, and triangular heatmap all refer to the same variant sequence.

### Step 3. Inspect the LDTrack object and pairwise statistics

`LDTrack` keeps the complete state of an LD calculation:

```{r}
names(ld_high)
```

The main components are:

| Component | Content | Typical use |
| --- | --- | --- |
| `data` | one row per unique variant pair | inspect `r2`, `Dprime`, distance, sample number, and allele frequencies |
| `matrix` | symmetric matrix of the selected LD metric | matrix-based inspection or export |
| `variants` | ordered metadata for retained variants | confirm positions and IDs |
| `region` | chromosome and queried boundaries | record the analyzed interval |
| `genotype` | optional variant-by-sample dosage matrix | detailed QC when explicitly retained |
| `figure` | figure created by `plot_ld_block()` | plotting/export |
| `meta` | method, sample count, ploidy, and calculation notes | reproduce the analysis settings |

Inspect the pairwise table directly:

```{r}
ld_high$data[, .(
  variant_i,
  variant_j,
  distance_bp,
  n_samples,
  r,
  r2,
  Dprime,
  ld
)]
```

Important columns are:

- `distance_bp`: physical distance between the two variants;
- `n_samples`: paired non-missing sample number used for that comparison;
- `r`: dosage correlation;
- `r2`: squared dosage correlation;
- `Dprime`: dosage-based D-prime approximation;
- `ld`: the metric selected by `method`; for the default `method = "r2"`, `ld` is identical to `r2`.

The region and settings used for calculation are preserved in the object:

```{r}
ld_high$region
ld_high$meta
```

### Step 4. Inspect the LD matrix and descriptive pair filters

The symmetric LD matrix uses variant IDs as row and column names and contains 1 on the diagonal:

```{r}
round(ld_high$matrix, 3)
```

For the designed high-LD region, every off-diagonal value is also 1.

GeneTrackR calculates pairwise LD but does **not** automatically call biological LD-block boundaries from a fixed threshold. If a project needs a descriptive high-LD pair set, filter the pairwise result explicitly and keep the chosen threshold visible in the analysis:

```{r}
strong_pairs <- ld_high$data[
  !is.na(r2) & r2 >= 0.8,
  .(variant_i, variant_j, distance_bp, r2)
]

strong_pairs
```

A threshold such as `r2 >= 0.8` is therefore a user-defined interpretation rule, not an implicit block-calling rule inside `compute_ld_block()`.

### Step 5. Draw the triangular LD heatmap

Plot the calculated object and store the returned updated `LDTrack`:

```{r}
ld_high <- plot_ld_block(
  ld_high,
  color_palette = "Paired",
  label_by = "variant_id",
  show_variant_labels = TRUE,
  show_region = FALSE,
  return_object = TRUE
)

ld_high$figure
```

`plot_ld_block()` returns the updated `LDTrack` by default. The calculation results remain unchanged and the generated ggplot/patchwork figure is stored in:

```{r}
ld_high$figure
```

This assignment pattern is recommended:

```text
LDTrack <- compute_ld_block(...)
LDTrack <- plot_ld_block(LDTrack, ...)
figure  <- LDTrack$figure
```

For dense regions, hide x-axis labels:

```{r}
ld_high_no_labels <- plot_ld_block(
  ld_high,
  show_variant_labels = FALSE,
  color_palette = "Paired",
  return_object = TRUE
)

ld_high_no_labels$figure
```

`label_by = "pos"` shows genomic positions; `label_by = "variant_id"` shows VCF IDs. Another column in `ld$variants` can also be supplied when present.

### Step 6. Add gene structure and natural-variant markers

For locus interpretation, add the compact genomic region track above the triangular heatmap:

```{r}
ld_high_region <- plot_ld_block(
  ld_high,
  show_region = TRUE,
  annotation = gp,
  connect_region = TRUE,
  show_variant_marker = TRUE,
  variant_marker_size = 3,
  label_by = "variant_id",
  show_variant_labels = TRUE,
  color_palette = "Paired",
  region_height = 1.25,
  connector_height = 0.35,
  return_object = TRUE
)

ld_high_region$figure
```

The combined plot contains three aligned components when `connect_region = TRUE`:

1. compact gene/variant region track;
2. connector lines from genomic variant positions to LD columns;
3. triangular LD heatmap.

Variant triangle markers can be hidden without removing the gene track:

```{r}
ld_high_region_no_markers <- plot_ld_block(
  ld_high,
  show_region = TRUE,
  annotation = gp,
  show_variant_marker = FALSE,
  show_variant_labels = FALSE,
  color_palette = "Paired",
  return_object = TRUE
)

ld_high_region_no_markers$figure
```

Alternatively, set `variant_marker_size = 0`. `region_height`, `connector_height`, and `heatmap_height` control the relative vertical space of the combined panels.

### Step 7. Compare high-LD, low-LD, and two-variant regions

A high-LD heatmap is easiest to interpret when contrasted with an intentionally weak-LD region. The demo low-LD interval contains four variants:

```{r}
ld_low <- compute_ld_block(
  vcf,
  chrom = "chr2",
  start = 1999000,
  end = 2008500,
  method = "r2",
  verbose = FALSE
)

ld_low$data[, .(
  variant_i,
  variant_j,
  distance_bp,
  r2
)]
```

The six pairwise values are designed so that most are zero and the maximum is approximately `1 / 9`:

```{r}
range(ld_low$data$r2, na.rm = TRUE)
max(ld_low$data$r2, na.rm = TRUE)
```

Plot the low-LD region with the same palette and scale as the high-LD example:

```{r}
ld_low <- plot_ld_block(
  ld_low,
  label_by = "variant_id",
  color_palette = "Paired",
  return_object = TRUE
)

ld_low$figure
```

The two-variant demo region contains exactly one pairwise comparison:

```{r}
ld_pair <- compute_ld_block(
  vcf,
  chrom = "chr2",
  start = 16995001,
  end = 17006000,
  method = "r2",
  verbose = FALSE
)

ld_pair$data
```

`plot_ld_block()` draws this case as one centered diamond rather than attempting to construct a larger triangular matrix:

```{r}
ld_pair <- plot_ld_block(
  ld_pair,
  label_by = "variant_id",
  color_palette = "Paired",
  return_object = TRUE
)

ld_pair$figure
```

When only the figure is needed for a temporary plotting call, compatibility mode can return it directly:

```{r}
pair_figure <- plot_ld_block(
  ld_pair,
  return_object = FALSE
)

pair_figure
```

The standard workflow should still retain the `LDTrack` object because it preserves the calculation and figure together.

### Step 8. Control samples, variant types, LD methods, and genotype retention

#### Select samples

Use `samples` when LD should be estimated in a defined population subset:

```{r}
subset_samples <- sprintf("S%02d", 1:24)

ld_subset <- compute_ld_block(
  vcf,
  chrom = "chr1",
  start = 12342620,
  end = 12343180,
  samples = subset_samples,
  min_pair_samples = 10,
  method = "r2",
  verbose = FALSE
)

ld_subset$meta$sample_n
```

LD is population dependent. Results calculated in one sample subset should not be interpreted as if they necessarily apply to another population.

#### Select variant types

`variant_type` controls the records retained before pairwise calculation:

```{r}
ld_snp <- compute_ld_block(
  vcf,
  chrom = "chr1",
  start = 12342620,
  end = 12343180,
  variant_type = "snp",
  method = "r2",
  verbose = FALSE
)
```

Available choices are:

- `both`: retain SNPs and indel/MNV records;
- `snp`: retain SNPs only;
- `ind`: retain INS/DEL/MNV/INDEL records.

At least two retained variants are required.

#### Retain the dosage matrix when needed

The genotype dosage matrix is omitted by default to reduce object size. Retain it only when detailed genotype QC or downstream matrix analysis is needed:

```{r}
ld_with_gt <- compute_ld_block(
  vcf,
  chrom = "chr1",
  start = 12342620,
  end = 12343180,
  method = "r2",
  keep_genotype_matrix = TRUE,
  verbose = FALSE
)

dim(ld_with_gt$genotype)
ld_with_gt$genotype[, 1:6, drop = FALSE]
```

The dosage definition is the number of non-reference allele copies parsed from `GT` for diploid biallelic records. Multi-allelic non-reference alleles are collapsed into total non-reference dosage.

#### Use D-prime cautiously

`method = "Dprime"` is also available:

```{r}
ld_dprime <- compute_ld_block(
  vcf,
  chrom = "chr2",
  start = 1999000,
  end = 2008500,
  method = "Dprime",
  verbose = FALSE
)

ld_dprime$data[, .(
  variant_i,
  variant_j,
  r2,
  Dprime,
  ld
)]
```

The current `Dprime` implementation is a **dosage-based approximation for unphased VCF genotypes**. It should not be described as an exact haplotype-based D-prime estimate. Use phased haplotypes when exact haplotype-frequency-based D-prime is required. For ordinary unphased genotype VCF workflows, `r2` is the recommended primary GeneTrackR LD metric.

### Step 9. Interpret LD and pass the selected region downstream

LD measures statistical association between variants; it does not by itself identify a causal variant or prove that all variants in a visually strong region should be treated as one biological unit. Interpret the LD pattern together with:

- the lead or focal variant;
- physical distance;
- gene structure and regulatory annotations;
- sample population;
- variant quality and missingness;
- phenotype or association evidence.

Once an interval has been selected biologically, the same coordinates can be passed directly to the haplotype workflow:

```{r}
hap_ld_region <- hap_region_variant(
  vcf,
  chrom = "chr1",
  start = 12342620,
  end = 12343180,
  genotype_mode = "string"
)

hap_ld_region
```

This keeps the analysis chain explicit:

```text
VCF genotypes
    -> pairwise LD calculation
    -> inspect LDTrack and regional gene context
    -> define the biological interval of interest
    -> construct regional haplotypes
    -> phenotype/refinement analysis if appropriate
```

For real GWAS loci, it is usually preferable to define the LD query around a lead SNP or candidate interval first, inspect the pairwise pattern, and only then decide which linked variants or genes should move into the haplotype analysis.
