## Linkage disequilibrium analysis

The demo VCF contains a designed six-variant high-LD block (`varLD01`-`varLD06`) in `GeneA`. These variants share the same genotype pattern and therefore provide a deterministic `r2 = 1` example.

```{r}
ld <- compute_ld_block(
  vcf,
  chrom = "chr1",
  start = 12342620,
  end = 12343180,
  method = "r2",
  verbose = FALSE
)

ld$data
```

Add the compact GeneA structure above the triangular heatmap:

```{r}
ld <- plot_ld_block(
  ld,
  show_region = TRUE,
  annotation = gp,
  show_variant_labels = FALSE
)

ld$figure
```

The assignment is intentional: `plot_ld_block()` returns the updated `LDTrack`, not the figure, when `return_object = TRUE` (the default).

The `GeneT` region contains exactly two variants and is retained as the deterministic two-variant LD plotting case:

```{r}
ld_pair <- compute_ld_block(
  vcf,
  chrom = "chr2",
  start = 16995001,
  end = 17006000,
  method = "r2",
  verbose = FALSE
)

plot_ld_block(ld_pair, return_object = FALSE)
```
